### 2014-10-22 retention.tex
init

\documentclass[a4paper]{article}
\usepackage{graphicx}
\usepackage{wrapfig}

\title{Peptide retention time prediction}

\author{
Luminita Moruz\\
Science for Life Laboratory,\\
School of Biotechnology,\\
Royal Institute of Technology (KTH),\\
Stockholm, Sweden
\and
Lukas K\"all\\
Science for Life Laboratory,\\
School of Biotechnology,\\
Royal Institute of Technology (KTH),\\
Stockholm, Sweden}

\begin{document}

\maketitle

\setcounter{secnumdepth}{2} % organisational level that receives a numbers
\setcounter{tocdepth}{2}    % print table of contents for level 2
\tableofcontents            % print the table of contents
% levels are: 0 - chapter, 1 - section, 2 - subsection, 3 - subsubsection

\section{Introduction}

Besides decreasing the number of peptides injected at the same time
into the mass spectrometer, the separation process in itself provides
useful information about the analytes. On the one hand, the instrument
always records the time when a peptide elutes from the chromatography
column, i.e. the retention time. On the other hand, the retention time
can also be estimated by examining the sequence of the peptide. The
availability of both predicted and observed retention times has proven
to be a powerful feature, with several applications in mass
spectrometry-based proteomics \cite{Palmblad2013}.  In this chapter,
Sections~\ref{sec:rtpred} and \ref{sec:rtpredm} describe the main
approaches used to estimate retention time from peptide sequences,
while Section~\ref{sec:app} discusses some of the most important
applications of retention time predictions. Section~\ref{sec:irt}
briefly introduces the use of peptide retention standards.

\section{\label{sec:rtpred}Retention time prediction for unmodified peptides}

\begin{wrapfigure}{r}{.45\textwidth}
\vspace{-10pt}
%\begin{figure}[htbp]
\centering
\includegraphics[trim=0.5cm 0cm 2cm 1.5cm, clip=true, width=0.43\textwidth]{img/reproducibility.pdf}
\caption{\label{fig:repr} {\bf Reproducibility of peptide retention time.} Two identical shotgun proteomics experiments were carried out. The retention times of the peptides confidently identified in both runs are displayed.}
%\end{figure}
\vspace{-12pt}
\end{wrapfigure}

Liquid chromatography gives reproducible results. Except for small
shifts from run to run, the same peptide elutes at about the same time
when a sample is reanalyzed under identical experimental conditions
(Figure~\ref{fig:repr}). This suggests that peptides engage in
well-defined interactions during separation, which depend only on the
experimental setup employed and the properties of the peptides
themselves. However, the properties of the peptides are determined to
a large extent by their amino-acid sequence. As a consequence, given a
chromatography setup, the sequence of a peptide should provide
virtually all the information needed to predict its retention time.

\vspace{0.15cm}

Following these observations, a variety of retention time prediction
methods has been proposed. These methods can be roughly divided into
three classes: index-based methods, modeling-based techniques and
machine learning approaches. The main ideas underlying each category
are described below.

\subsection{Index-based methods}
\hyphenation{lear-ning}
\hyphenation{indivi-dual}

% Basic idea 
Index-based prediction methods aim at estimating the effect that each
individual amino acid in a sequence has on the retention time of the
peptide. The individual contributions of the amino acids are often
referred to as retention coefficients, and a set of retention
coefficients forms a retention index. Once a retention index is
derived for a given chromatography system, the retention time of a
peptide can be estimated by simply summing up the retention
coefficients of the constituent amino acids.

\vspace{0.15cm}

% The first method 
The first method to implement this idea used a set of 25 short
peptides together with their observed retention times to derive
retention coefficients for each of the amino acids and end groups
present in the sequences \cite{meek1980}. The coefficients were
calculated using a form of iterative regression and were estimated
using only the amino-acid composition of the peptides, without any
information about the position of each amino acid in the
sequence. However, since shuffled versions of the same peptide
sequence had previously been separated on a RPLC column, the author
pointed out that additional parameters would be required for
predicting the retention time of longer peptides.

%The study reported good correlations between predicted and observed
%retention times, showing that for short peptides the amino acid
%composition provided sufficient information to produce reasonable
%retention time predictions.

\vspace{0.15cm}

% It became available that additional information is needed; anc oclumn is important  examples 
Indeed, a subsequent study showed that even for short peptides,
distinct retention coefficients are obtained when an amino acid is
located at different positions in a sequence \cite{Houghten1987}. In
this context, the authors speculated that different retention
coefficients would be needed for each position in a peptide
sequence. Although the exact chemical mechanisms underlying this
observation are still not fully understood, a number of studies have
pinpointed individual factors that alter the retention coefficients of
amino acids. As an example, it has been shown that the retention
contribution of an amino acid is affected by the overall
hydrophobicity of the peptide \cite{Mant2006}, the number of positive
charges \cite{Mant2006} or presence of elements of secondary
structure \cite{Zhou1990}. In addition, modifications in
chromatography setup such as changing the composition of the mobile or
stationary phases also modify the elution pattern of the
peptides \cite{Browne1982, Guo1987, Gilar2010}.

\vspace{0.15cm}
% SSRcalc was proposed including a number of additional terms; then it was extended
% for a few other column
Following these observations, Krokhin et al. developed
SSRCalc \cite{Krokhin2004}, which is currently the most widely used
index-based retention time predictor. In the initial version of the
algorithm a collection of 346 tryptic peptides was used to derive two
sets of retention coefficients, one corresponding to the N-terminal
and one to all the other positions in a peptide sequence. Furthermore,
correction factors for peptide length and overall hydrophobicity were
included. Subsequent versions of the algorithm included terms
accounting for additional properties such as isoelectric point,
nearest-neighbor effects of charged peptides and propensity to form
helical structures \cite{Krokhin2006}. However, a limitation of the
approach is that the algorithm was optimized for predicting retention
times for certain chromatography setups. Thus, SSRCalc's prediction
ability deteriorates for experimental conditions diverging from the
setup on which the algorithm was optimized \cite{Spicer2007}.

% - sum of retention coefficients (Meek reference 22) - different RT
% coefficients and the observation that this coefficients depend on
% the chromatography system (Browne ref 24) - investogate the
% contribution of individual amino acids (Guo 26,27) - all the work
% above was done on synthetic peptides os simple samples. Then RT
% prediction for complex sa% mples was proposed by (Palmblad 28-30) -
% important the exact position of each aa (pref 27) - additional rules
% included peptide length (ref 32), position of K and R residues [ref
% 33], and in the e% nd that the position of each amino acid at each
% position is important [ref 34] - amphipathicity [ref 35] - SSRCalc
% [ref 36-39]

\subsection{Modeling-based approaches}

Modeling-based techniques predict retention times by using information
about the structure of the peptides and/or their chemical interactions
during separation. An example is the methods based on Quantitative
Structure-Retention Relationships (QSRR) \cite{Kaliszan2005,
Baczek2005}. In such approaches specialized molecular modeling
software is used to calculate a number of chemical descriptors from
the peptide sequence. The most relevant such descriptors are then
selected and combined via regression analysis into a prediction
function, which is subsequently used to estimate retention time for
other peptides of interest.

\vspace{0.15cm}

In other modeling-based approaches theoretical models are used to
capture the chemical behavior of peptides on the chromatography
column. One such method is BioLCCC, which attempts to model the
connectivity of the amino acids in a peptide chain and the general
mechanisms of chromatographic
separation \cite{gorshkov2006}. Nevertheless, given the numerous
factors involved in peptide separation, modeling approaches often
involve a considerable number of simplifications.

\subsection{Machine learning-based methods}

In machine learning-based approaches a set of training peptides,
i.e. peptide sequences with known retention times, are used to
customize a predefined mathematical model. This step is typically
called training or learning. Once trained, the mathematical model can
be subsequently used to predict retention time for other peptides of
interest.

\vspace{0.15cm}

One of the first machine learning approaches for retention time
prediction was based on an Artificial Neural Network
(ANN) \cite{petritis2003}. Initially, each peptide was described by 20
features giving the number of each of the 20 amino acid residues
present in the sequence. The network was trained using $\sim$7000
peptides with known retention times and evaluated using a set of
$\sim$5000 peptides originating from another organism. The algorithm
reported good accuracies, especially taking into account that it only
used information about the amino acid composition of the peptides.

\vspace{0.15cm}

In a follow-up work, the method was extended by representing each
 peptide by 1052 instead of 20
 features \cite{petritis2006improved}. The new descriptors encoded the
 identity of each amino acid at each position in a sequence, as well
 as the length and the hydrophobic moment of the peptide. The new
 neural network was trained on more than 300,000 peptide sequences. To
 obtain such a large number of peptides with known retention times, a
 strategy to normalize elution times was developed. This allowed
 pooling of peptides identified in more than 12,000 shotgun
 experiments. The trained algorithm reported excellent performance for
 the 1303 peptides on which it was evaluated.

\vspace{0.15cm}

A drawback of the method was however the large amount of training
peptides required, which made the algorithm difficult to retrain for
other chromatographic conditions. This was a serious limitation in the
context of RPLC separation, where a broad variety of chromatographic
setups is commonly used across research laboratories. To address this,
Klammer et al. proposed a predictor based on support vector machines
(SVMs) \cite{klammer2007improving}. Their method required fewer
training peptides and could thus be easily adapted to new
chromatographic conditions, but yielded lower prediction
accuracies. Subsequently, a more elaborate SVM-based method using a
customized kernel function was proposed \cite{rtpredictImproved}. This
approach provided reasonable prediction accuracy, while requiring as
little as 50 peptides for training.

% Other retention time prediction method used a feature selection
% approach to select relevant descriptors, and showed good
% performances even for longer peptides containing up to 50 amino
% acids \cite{shinoda2006}.

\vspace{0.15cm}

However, despite all these efforts, it should be pointed out that none
of the retention time prediction methods currently available is able
to provide accuracies comparable to the small shifts observed when
running a sample twice through a chromatography column
(Figure \ref{fig:repr}). Recently, the lack of suitable
representations for secondary structure elements was indicated as one
of the main causes of prediction errors \cite{Reimer2012}.  Most
probably additional factors are involved, and further endeavors are
required to reach the desired level of accuracy.

\section{\label{sec:rtpredm}Retention time prediction for modified peptides}

% very few studies have documented the chromatographic behavior of
% modified peptides
The study of peptide separation mechanisms in RPLC has been an active
area of research for the past 30 years. Nonetheless, surprisingly
little efforts have been directed toward investigating the
chromatographic behavior of posttranslationally modified peptides. The
few studies available so far have mostly focused on phosphorylated
peptides \cite{Kim2007}, or a couple of artifactual modifications
commonly introduced during sample
preparation \cite{Reimer2011}. Consequently, few retention time
prediction methods have been extended and assessed for such
peptides. A majority of the approaches currently available are
modification-specific, many of them developed specifically for
phosphopeptides \cite{Kawakami2005, perlova2010}.

\vspace{0.15cm}

As large-scale studies of a variety of posttranslational modifications
increasingly come within reach (Section \ref{sec:ptms}), predictors
handling other modifications will become available. In this context,
machine learning methods are expected to play an important role
because of their flexibility and intrinsic ability to adapt to new
datasets.

\section{\label{sec:app}Applications of retention time prediction}

% RT prediction is used to increase number of identifications 
The time points when peptides elute from the RPLC column are typically
recorded in MS experiments. By comparing these values with predicted
retention times for peptides from a sequence database, information
about the identity of the peptides can be obtained. This relatively
straightforward idea has been implemented in a variety of flavors and
was proven to bring significant improvements in the peptide
identification process.

\vspace{0.15cm}

Palmblad et al. showed that matching theoretical and observed masses
and retention times provides a simple, yet powerful method to identify
peptides in simple biological
mixtures \cite{palmblad2002prediction}. Strittmatter et al. used
retention time as a parameter in a discriminant function that
differentiated between correct and incorrect peptide
identifications \cite{Strittmatter2004}. Following this approach, they
reported $\sim$9\% more confident peptide identifications in a typical
shotgun experiment. Along the same lines, Klammer et al. demonstrated
that for specific experimental conditions, just filtering out peptides
with large deviations between observed and predicted retention time
leads to up to 50\% more
identifications \cite{klammer2007improving}.

\vspace{0.15cm}

Krokhin et al. illustrated yet another route to employ retention time
predictions \cite{krokhin200612, McQueen2012}. They showed that
theoretical retention times and mass measurements can be used to
define lists of peptides belonging to already identified proteins. The
instrument can be instructed to avoid fragmenting these peptides, thus
decreasing the number of fragmentation events that need to be acquired
for identifying a set of proteins.

\vspace{0.15cm}

% Improve the design of SRM experiments 
In addition, predicted retention times can also be used to investigate
the elution properties of the peptides, and possibly design improved
experimental setups. As an example, Bertsch et al. employed retention
time predictions to design {\em de novo} targeted proteomic
assays \cite{bertsch2010}. Other applications include decreasing the
number of candidate peptides during database
searching \cite{lobas2013} and simulations of MS-based
experiments \cite{bielow2011}.

% It is an orthogonal information. Decrease the database search space,
% study properties of the peptides

\section{\label{sec:irt}Peptide retention standards}

\hyphenation{wide-spread}
% peptide standards what they are, what they are used for 
In the past years, the use of retention standards has become
increasingly widespread in the context of mass spectrometry-based
proteomics~\cite{olegstd, irt}. Such standards typically consist of a
small number of peptides covering a wide span of hydrophobicities and
with well studied behaviors under LC-MS/MS. These peptides are often
analyzed together with the sample of interest, and by comparing their
observed and expected behaviors one can obtain valuable information
for a variety of applications.

\vspace{0.15cm}

As an example, in targeted proteomics peptide standards can be used to
 calibrate predicted retention times and to estimate the accuracy of
 the prediction algorithms~\cite{Kiyonami01022011}. In addition,
 recent studies have reported excellent results when using peptide
 standards to adjust the time when certain transitions are scheduled
 ``on-the-fly''~\cite{seb, irt}. Retention standards are also useful
 for aligning data from multiple experiments~\cite{petritis2003}, to
 transfer retention times between different chromatography
 setups~\cite{seb}, or to keep track of the performance of LC-MS/MS
 systems~\cite{qcal}.

\bibliographystyle{unsrt}
\bibliography{./retention.bib}

\end{document}

### 2014-11-08 retention.tex
Added a few paragraphs in the introduction

@@
 \section{Introduction}
+Mass spectrometry-based methods are unquestionably the most popular techniques for studying the proteint content of biological samples.
+The palette of protemic applications where such methods are used is getting increasingly diversified, ranging from identifying and quantifying
+proteins in biological samples \cite{Geiger2012} to determining
+posttranslational modifications \cite{Huttlin2010} and elucidating
+protein interaction networks \cite{Gavin2011}.
 
-Besides decreasing the number of peptides injected at the same time
-into the mass spectrometer, the separation process in itself provides
+Although available in a variety of flavours, most of the widely spread MS-based workflows consists of a few common steps.
+First, the proteins in the mixture of interest are cleaved into peptides using a sequence-specific enzyme such as trypsin.
+This step is necessary because the mass spectrometers are more sensitive for peptides, and because spectra produced by full-length proteins are difficult to interpret.
+In the next step, the resulting peptides are separated according to their hydrophobicity on a liquid chromatography (LC) column.
+As they gradually elute from the column, the peptides are ionized via electrospray ionization and subjected to mass spectrometry analysis.
+The result of these measurements consists of a set of spectra, which are subsequently used to infer valuable information about the proteins present in the initial sample.
+
+Today a majority of the mass spectrometry (MS) protocols in use are high-throughput, generating extensive amounts of data.
+This has brought up a whole new set of analytical challenges, making the development of efficient algorithms to manipulate and interpret the data a major area of proteomics research.
+A particular difficult challenge in this context is posed by the tremendous complexity of the data.
+Proteins
+display a wide range of physico-chemical properties and are often found at very different amounts in complex samples, spanning as much as 12 orders of magnitude in relative
+abundance~\cite{Angel2012}. Furthermore, a large proportion of
+proteins are subjected to posttranslational modifications. As an
+example, a few years ago more than 25,000 phosphorylation sites on
+human proteins were already reported~\cite{Lemeer2009}. To manage this formidable diversity,
+ increasing efforts are made to extract useful information not only from the spectra, but also from other types of information recorded during MS experiments.
+
+An example of such information is retention time, i.e.
+the time point when a peptide elutes from the liquid chromatography column.
+Although the main purpose of liquid chromatography is to reduce the number of peptides
+simultaneously injected into the mass spectrometer, and by this
+increase the proteome coverage achieved in an experiment. However, besides this, the separation process in itself provides
 useful information about the analytes. On the one hand, the instrument
@@
 to be a powerful feature, with several applications in mass
-spectrometry-based proteomics \cite{Palmblad2013}.  In this chapter,
+spectrometry-based proteomics \cite{Palmblad2013}.
+
+In this review we summarize the work done in the field of retention time prediction, and describe a few of the successful applications of such predictions. 
 Sections~\ref{sec:rtpred} and \ref{sec:rtpredm} describe the main
-approaches used to estimate retention time from peptide sequences,
-while Section~\ref{sec:app} discusses some of the most important
+approaches used to estimate retention time from peptide sequences.
+Section~\ref{sec:app} discusses some of the most important
 applications of retention time predictions. Section~\ref{sec:irt}

### 2014-11-08 retention.tex
Added a paragraph about Elude and a table listing rt methods

@@
 \usepackage{wrapfig}
+\usepackage{url}
+\usepackage{multirow}
 
@@
 
+Liquid chromatography gives reproducible results. Except for small
+shifts from run to run, the same peptide elutes at about the same time
+when a sample is reanalyzed under identical experimental conditions
+(Figure~\ref{fig:repr}). This suggests that peptides engage in
+well-defined interactions during separation, which depend only on the
+experimental setup employed and the properties of the peptides
+themselves. However, the properties of the peptides are determined to
+a large extent by their amino-acid sequence. As a consequence, given a
+chromatography setup, the sequence of a peptide should provide
+virtually all the information needed to predict its retention time.
+
 \begin{wrapfigure}{r}{.45\textwidth}
@@
 
-Liquid chromatography gives reproducible results. Except for small
-shifts from run to run, the same peptide elutes at about the same time
-when a sample is reanalyzed under identical experimental conditions
-(Figure~\ref{fig:repr}). This suggests that peptides engage in
-well-defined interactions during separation, which depend only on the
-experimental setup employed and the properties of the peptides
-themselves. However, the properties of the peptides are determined to
-a large extent by their amino-acid sequence. As a consequence, given a
-chromatography setup, the sequence of a peptide should provide
-virtually all the information needed to predict its retention time.
-
-\vspace{0.15cm}
-
 Following these observations, a variety of retention time prediction
@@
 machine learning approaches. The main ideas underlying each category
-are described below.
+are described below, while Table \ref{tab:rtmethods} lists a few of the most popular retention time 
+prediction methods from each category.
 
@@
 
+In our group we developed a software package which we named {\sc
+Elude} \cite{elude1, elude2}. One of the main novelty
+points of {\sc Elude} is the calculation of peptide features based on
+both a general hydrophobicity index, as well as a retention index
+trained for each dataset at hand. This ensures that the algorithm
+can be applied to different chromatography setups with minimal loss
+in performance. For cases when little training data is available,
+{\sc ELUDE} implements an additional workflow based on a library of
+pre-trained retention models. This workflow requires only a few tens
+of training peptides to select and calibrate the most suitable model
+from the library. Both the selection and calibration are performed
+using statistical methods robust to outliers.
+
+\vspace{0.15cm}
+
 However, despite all these efforts, it should be pointed out that none
@@
 required to reach the desired level of accuracy.
+
+\begin{table}
+ \caption{{\bf Retention time prediction methods.} We
+   give a few of the most commonly used retention time prediction methods available.}
+ \vspace{.2cm}
+ \label{tab:rtmethods}
+ \begin{tabular}{lll}
+
+ Method name & Link & Reference  \\
+ \hline
+SSRCalc & \url{http://hs2.proteome.ca/SSRCalc/SSRCalcX.html} & [\cite{Krokhin2004}, \cite{Krokhin2006}, \cite{Spicer2007}] \\
+BioLCCC & \url{http://theorchromo.ru/} & [\cite{gorshkov2006}] \\
+ANN & Link? & [\cite{petritis2003}, \cite{petritis2006improved}] \\
+RTModel, RTPredict &  \url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTPredict.html}, 
+ & [\cite{rtpredict}, \cite{rtpredictImproved}] \\
+&\url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTModel.html}  & \\
+
+Elude & \url{http://elude.sbc.su.se/} & [\cite{elude1}, \cite{elude2}] \\
+\hline
+ \end{tabular} \\
+\end{table}
 

### 2014-11-09 retention.tex
Added paragraph and image about elude-ptm

--- 2014-11-08 retention.tex
@@
 \documentclass[a4paper]{article}
+\usepackage[margin=0.5in]{geometry} 
 \usepackage{graphicx}
 \usepackage{wrapfig}
+\usepackage{url}
+\usepackage{multirow}
 
@@
 
-Although available in a variety of flavours, most of the widely spread MS-based workflows consists of a few common steps.
+\vspace{0.15cm}
+
+Although available in a variety of flavours, most of the widely spread MS-based workflows consists of the same main steps.
 First, the proteins in the mixture of interest are cleaved into peptides using a sequence-specific enzyme such as trypsin.
@@
 
+\vspace{0.15cm}
+
 Today a majority of the mass spectrometry (MS) protocols in use are high-throughput, generating extensive amounts of data.
-This has brought up a whole new set of analytical challenges, making the development of efficient algorithms to manipulate and interpret the data a major area of proteomics research.
+This has brought up a whole new set of analytical challenges, making the development of efficient algorithms to interpret this data a major area of proteomics research.
 A particular difficult challenge in this context is posed by the tremendous complexity of the data.
@@
  increasing efforts are made to extract useful information not only from the spectra, but also from other types of information recorded during MS experiments.
-
 An example of such information is retention time, i.e.
 the time point when a peptide elutes from the liquid chromatography column.
-Although the main purpose of liquid chromatography is to reduce the number of peptides
+
+\vspace{0.15cm}
+
+The main purpose of liquid chromatography is to reduce the number of peptides
 simultaneously injected into the mass spectrometer, and by this
@@
 
+\vspace{0.15cm}
+
 In this review we summarize the work done in the field of retention time prediction, and describe a few of the successful applications of such predictions. 
@@
 \section{\label{sec:rtpred}Retention time prediction for unmodified peptides}
-
-\begin{wrapfigure}{r}{.45\textwidth}
-\vspace{-10pt}
-%\begin{figure}[htbp]
-\centering
-\includegraphics[trim=0.5cm 0cm 2cm 1.5cm, clip=true, width=0.43\textwidth]{img/reproducibility.pdf}
-\caption{\label{fig:repr} {\bf Reproducibility of peptide retention time.} Two identical shotgun proteomics experiments were carried out.
-The retention times of the peptides confidently identified in both runs are displayed.}
-%\end{figure}
-\vspace{-12pt}
-\end{wrapfigure}
 
@@
 
+\begin{wrapfigure}{r}{.45\textwidth}
+\vspace{-10pt}
+%\begin{figure}[htbp]
+\centering
+\includegraphics[trim=0.5cm 0cm 2cm 1.5cm, clip=true, width=0.43\textwidth]{img/reproducibility.pdf}
+\caption{\label{fig:repr} {\bf Reproducibility of peptide retention time.} Two identical shotgun proteomics experiments were carried out.
+The retention times of the peptides confidently identified in both runs are displayed.}
+%\end{figure}
+\vspace{-25pt}
+\end{wrapfigure}
+
 \vspace{0.15cm}
@@
 predicting the retention time of longer peptides.
+
+\begin{table}
+ \caption{{\bf Retention time prediction methods.} We
+   give a few of the most commonly used retention time prediction methods available.}
+ \vspace{.2cm}
+ \label{tab:rtmethods}
+ \begin{tabular}{lll}
+
+ Method name & Link & References  \\
+ \hline
+SSRCalc & \url{http://hs2.proteome.ca/SSRCalc/SSRCalcX.html} & \cite{Krokhin2004, Krokhin2006}, \cite{Spicer2007} \\
+BioLCCC & \url{http://theorchromo.ru/} & \cite{gorshkov2006} \\
+ANN & Link? & \cite{petritis2003 ,petritis2006improved} \\
+RTModel, RTPredict &  \url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTPredict.html}, 
+ & \cite{rtpredict, rtpredictImproved} \\
+&\url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTModel.html}  & \\
+
+Elude & \url{http://elude.sbc.su.se/} & \cite{elude1, elude2} \\
+\hline
+ \end{tabular} \\
+\end{table}
 
@@
 accuracies. Subsequently, a more elaborate SVM-based method using a
-customized kernel function was proposed \cite{rtpredictImproved}. This
+customized kernel function was proposed \cite{rtpredict, rtpredictImproved}. This
 approach provided reasonable prediction accuracy, while requiring as
@@
 % acids \cite{shinoda2006}.
+
+\vspace{0.15cm}
+
+In our group we developed a software package for retention time prediction which we named {\sc
+Elude} \cite{elude1, elude2}. One of the main novelty
+points of {\sc Elude} wass the calculation of peptide features based on
+both a general hydrophobicity index, as well as a retention index
+trained for each dataset at hand. This ensured that the algorithm
+could be applied to different chromatography setups with minimal loss
+in performance. For cases when little training data was available,
+{\sc ELUDE} implemented an additional workflow based on a library of
+pre-trained retention models. This workflow required only a few tens
+of training peptides to select and calibrate the most suitable model
+from the library. Both the selection and calibration were performed
+using statistical methods robust to outliers.
 
@@
 phosphopeptides \cite{Kawakami2005, perlova2010}.
+
+\begin{figure}[!h]
+\centering 
+\includegraphics[width=0.9\textwidth]{img/elude-ptm.pdf}
+\caption{\label{fig:elude} {\bf {\sc Elude}'s workflow.}  {\sc Elude} encompasses two learning steps.
+In the first one the training peptides are used to build an in-house retention index.
+This index, together with additional information, is then employed to calculate a set of 60 numerical features for each peptide.
+These features are optimally combined during a second training procedure.
+The resulting retention model can be subsequently used to predict the retention time of other peptides of interest.}
+\end{figure}
+
+\vspace{0.15cm}
+
+To address this, in our group we extended the first version of our algorithm {\sc ELUDE} to encompass modified peptides as well.
+This was straightforward to achieve once we observed that the in-house trained  retention index contained nearly all the information incorporated in the more general hydrophobicity scale.
+This meant that we could remove all the features based on this index with minimal loss in accuracy.
+Further, we redefined all features describing a peptide in terms of the in-house retention index alone.
+Since this index was trained for each dataset at hand, the algorithm became fully trainable for any modified amino acid (Figure \ref{fig:elude}).
+We evaluated the new version of {\sc ELUDE} on peptides containing artifactual modifications introduced as part of a MS protocol called COmbined
+FRActional DIagonal Chromatography (COFRADIC) \cite{Gevaert2002}. Our results showed that  {\sc ELUDE} yielded equally accurate predictions for modified and
+unmodified peptides  \cite{elude2}.
 

### 2014-11-09 retention.tex
Added some paragraphs about gradient optimization

@@
 Since this index was trained for each dataset at hand, the algorithm became fully trainable for any modified amino acid (Figure \ref{fig:elude}).
-We evaluated the new version of {\sc ELUDE} on peptides containing artifactual modifications introduced as part of a MS protocol called COmbined
+We evaluated the new version of {\sc ELUDE} on peptides containing artifactual modifications introduced as part of an MS protocol called COmbined
 FRActional DIagonal Chromatography (COFRADIC) \cite{Gevaert2002}. Our results showed that  {\sc ELUDE} yielded equally accurate predictions for modified and
@@
 As large-scale studies of a variety of posttranslational modifications
-increasingly come within reach (Section \ref{sec:ptms}), predictors
+increasingly come within reach, predictors
 handling other modifications will become available. In this context,
@@
 % Improve the design of SRM experiments 
-In addition, predicted retention times can also be used to investigate
-the elution properties of the peptides, and possibly design improved
-experimental setups. As an example, Bertsch et al. employed retention
-time predictions to design {\em de novo} targeted proteomic
-assays \cite{bertsch2010}. Other applications include decreasing the
+In addition, predicted retention times are useful investigating
+the elution properties of the peptides.  To illustrate this, we predicted the retention time of all the
+theoretical peptides from an {\it in silico} digest of the human
+proteome using  {\sc ELUDE}  \cite{elude1}. This exercise revealed a rather skewed
+distribution, with a majority of the peptides eluting in only a small
+portion of the chromatography run. This suggested that
+the linear gradient typically used in liquid chromatography are sub-optimal, and that one should be able to replace such gradients with nonlinear functions that produce even
+distributions of the peptides across runs.
+
+\vspace{0.15cm}
+
+We tested this hypothesis by calculating and evaluating
+two types of nonlinear gradients \cite{gradopt1}. Our results showed that the new
+gradients not only provided more uniform distributions of the
+peptides, but also led to an increase in the number of peptide
+identifications. In addition, they facilitated the identification of
+different peptide species compared to the typical linear gradients.
+To allow the easy calculation of nonlinear gradients by any MS practitioner, we also implemented a user-friendly graphical user interface that can be downloaded free of cost \cite{gradopt2}.
+Seen from a general perspective, this showed that the gradient function is yet another parameter that researchers can alter
+in their experiments, opening up a whole new range of potential
+applications.
+
+\vspace{0.15cm}
+
+Other examples of applications of retention time predictions include the optimization of  {\em de novo} targeted proteomic
+assays \cite{bertsch2010}, as well as decreasing the
 number of candidate peptides during database
-searching \cite{lobas2013} and simulations of MS-based
+searching \cite{lobas2013} and accurate simulations of MS-based
 experiments \cite{bielow2011}.
+
+%\begin{figure}[!h]
+%\centering 
+%\includegraphics[width=0.8\textwidth]{grad.pdf}
+%\caption{\label{fig:svm} {\bf Workflow to calculate nonlinear gradients.} We calculated two types of nonlinear gradients: one that evens out the predicted distribution of the peptides from an {\it in silico} digest of the proteome, and one that uniformizes the distribution of the abundant MS1-ions.}
+%\end{figure}
+
+\vspace{0.15cm}
+
+% what is the idea 
+%In Paper IV we tested this hypothesis by calculating and evaluating
+%two types of nonlinear gradients. Our results showed that the new
+%gradients not only provided more uniform distributions of the
+%peptides, but also led to an increase in the number of peptide
+%identifications. In addition, they facilitated the identification of
+%different peptide species compared to the typical linear gradients.
+%The project was carried out in collaboration with the Mechtler group
+%at the Research Institute of Molecular Pathology, Vienna.
 

### 2014-11-09 retention.tex
Added paragraphs about mf

@@
 
+In our group we have also investigated the possibility to use predicted retention times to increase the throughput of proteomics experiments.
+Our starting point was the observation that in typical shotgun proteomics experiments only a small fraction of the detected ions is
+fragmented, and thus used to identify
+peptides. The remaining unfragmented ions are usually disregarded in
+the subsequent data analyses. In the light of these observations, we investigated whether we could use
+information encapsulated by unfragmented ions to improve the protein
+inference in shotgun proteomics \cite{mf}.
+
+\vspace{0.15cm}
+
+% how it works and main results
+More specifically, we matched observed masses and retention times of
+unfragmented ions to predicted masses and retention times for the
+theoretical peptides from an {\it in silico} digest. We built a
+statistical framework to assign protein probabilities based on these
+matches, and combined the results with the ones obtained by the
+typical workflow based on fragmented ions. Following this approach, we
+obtained up to 7\% more protein identifications at a fixed false
+discovery rate threshold of 1\% \cite{mf}.
+
+\vspace{0.15cm}
+
+The idea of identifying proteins based on entire peptides, rather than
+peptide fragments, is not new. In fact, this was the principle
+underlying peptide mass fingerprinting, one of the first methods used
+to identify proteins via mass spectrometry. However, when introduced
+in the 90's, mass fingerprinting could not be applied to unequivocally
+identify proteins in complex biological samples. Consequently, shotgun
+proteomics has gained popularity instead. Nevertheless, with great
+developments in instrumentation leading to sub-ppm mass accuracies,
+the possibility of using mass fingerprinting-like methods came back to
+attention~\cite{iamt1}. Our study showed how improved mass accuracies
+and retention time predictions enable the use of unfragmented ions
+even in the context of complex peptide mixtures.
+
+\vspace{0.15cm}
+
 Krokhin et al. illustrated yet another route to employ retention time
@@
 
-\vspace{0.15cm}
-
-% what is the idea 
-%In Paper IV we tested this hypothesis by calculating and evaluating
-%two types of nonlinear gradients. Our results showed that the new
-%gradients not only provided more uniform distributions of the
-%peptides, but also led to an increase in the number of peptide
-%identifications. In addition, they facilitated the identification of
-%different peptide species compared to the typical linear gradients.
-%The project was carried out in collaboration with the Mechtler group
-%at the Research Institute of Molecular Pathology, Vienna.
-
 % It is an orthogonal information. Decrease the database search space,

### 2014-11-19 retention.tex
first stab on a discussion

@@
 \documentclass[a4paper]{article}
-\usepackage[margin=0.5in]{geometry} 
+\usepackage[margin=1in]{geometry} 
 \usepackage{graphicx}
@@
 
+\setlength{\parskip}{0.15cm}
+
 \section{Introduction}
-Mass spectrometry-based methods are unquestionably the most popular techniques for studying the proteint content of biological samples.
-The palette of protemic applications where such methods are used is getting increasingly diversified, ranging from identifying and quantifying
-proteins in biological samples \cite{Geiger2012} to determining
-posttranslational modifications \cite{Huttlin2010} and elucidating
-protein interaction networks \cite{Gavin2011}.
 
-\vspace{0.15cm}
+Mass spectrometry-based methods are unquestionably the most popular
+techniques for studying the proteint content of biological samples.
+The palette of protemic applications where such methods are used is
+getting increasingly diversified, ranging from identifying and
+quantifying proteins in biological samples \cite{Geiger2012} to
+determining posttranslational modifications \cite{Huttlin2010} and
+elucidating protein interaction networks \cite{Gavin2011}.
 
@@
 
-\vspace{0.15cm}
+Today a majority of the mass spectrometry (MS) protocols in use are
+high-throughput, generating extensive amounts of data. This has
+brought up a whole new set of analytical challenges, making the
+development of efficient algorithms to interpret this data a major
+area of proteomics research. A particular difficult challenge in this
+context is posed by the tremendous complexity of the data. Proteins
+display a wide range of physico-chemical properties and are often
+found at very different amounts in complex samples, spanning as much
+as 12 orders of magnitude in relative abundance~\cite{Angel2012}.
+Furthermore, a large proportion of proteins are subjected to
+posttranslational modifications. As an example, a few years ago more
+than 25,000 phosphorylation sites on human proteins were already
+reported~\cite{Lemeer2009}. To manage this formidable diversity,
+increasing efforts are made to extract as much information as possible
+from an experiment.
 
-Today a majority of the mass spectrometry (MS) protocols in use are high-throughput, generating extensive amounts of data.
-This has brought up a whole new set of analytical challenges, making the development of efficient algorithms to interpret this data a major area of proteomics research.
-A particular difficult challenge in this context is posed by the tremendous complexity of the data.
-Proteins
-display a wide range of physico-chemical properties and are often found at very different amounts in complex samples, spanning as much as 12 orders of magnitude in relative
-abundance~\cite{Angel2012}. Furthermore, a large proportion of
-proteins are subjected to posttranslational modifications. As an
-example, a few years ago more than 25,000 phosphorylation sites on
-human proteins were already reported~\cite{Lemeer2009}. To manage this formidable diversity,
- increasing efforts are made to extract useful information not only from the spectra, but also from other types of information recorded during MS experiments.
-An example of such information is retention time, i.e.
-the time point when a peptide elutes from the liquid chromatography column.
-
-\vspace{0.15cm}
-
-The main purpose of liquid chromatography is to reduce the number of peptides
-simultaneously injected into the mass spectrometer, and by this
-increase the proteome coverage achieved in an experiment. However, besides this, the separation process in itself provides
+The main purpose of liquid chromatography is to reduce the number of
+peptides simultaneously injected into the mass spectrometer, and by
+this increase the proteome coverage achieved in an experiment.
+However, besides this, the separation process in itself provides
 useful information about the analytes. On the one hand, the instrument
@@
 
-\vspace{0.15cm}
-
-In this review we summarize the work done in the field of retention time prediction, and describe a few of the successful applications of such predictions. 
-Sections~\ref{sec:rtpred} and \ref{sec:rtpredm} describe the main
-approaches used to estimate retention time from peptide sequences.
-Section~\ref{sec:app} discusses some of the most important
-applications of retention time predictions. Section~\ref{sec:irt}
-briefly introduces the use of peptide retention standards.
+In this review we summarize the work done in the field of retention
+time prediction, and describe a few of the successful applications of
+such predictions. Sections~\ref{sec:rtpred} and \ref{sec:rtpredm}
+describe the main approaches used to estimate retention time from
+peptide sequences. Section~\ref{sec:app} discusses some of the most
+important applications of retention time predictions.
+Section~\ref{sec:irt} briefly introduces the use of peptide retention
+standards.
 
@@
 
-\vspace{0.15cm}
+% iRT? \cite{Escher:2012qv}
+Following these observations, a variety of retention time prediction
+methods has been proposed. Most naturally one could
 
-Following these observations, a variety of retention time prediction
-methods has been proposed. These methods can be roughly divided into
-three classes: index-based methods, modeling-based techniques and
-machine learning approaches. The main ideas underlying each category
-are described below, while Table \ref{tab:rtmethods} lists a few of the most popular retention time 
-prediction methods from each category.
+These methods can be roughly divided into four classes: index-based
+methods, modeling-based techniques, machine learning approaches and
+look-up tables-based approaches. The main ideas underlying each
+category are described below, while Table \ref{tab:rtmethods} lists a
+few of the most popular retention time prediction methods from each
+category.
+
+\subsection{Methods based on look-ups of previous registered retention times}
+
+\cite{irt}
 
@@
 coefficients of the constituent amino acids.
-
-\vspace{0.15cm}
 
@@
 Elude & \url{http://elude.sbc.su.se/} & \cite{elude1, elude2} \\
+iRT-calculator & http://www.biognosys.ch/products/irt-calculator.html
+& \cite{irt} \\
 \hline
@@
 %retention time predictions.
-
-\vspace{0.15cm}
 
@@
 
-\vspace{0.15cm}
 % SSRcalc was proposed including a number of additional terms; then it was extended
@@
 
-\vspace{0.15cm}
-
 In other modeling-based approaches theoretical models are used to
@@
 
-\vspace{0.15cm}
-
 One of the first machine learning approaches for retention time
@@
 used information about the amino acid composition of the peptides.
-
-\vspace{0.15cm}
 
@@
  the 1303 peptides on which it was evaluated.
-
-\vspace{0.15cm}
 
@@
 
-\vspace{0.15cm}
-
 In our group we developed a software package for retention time prediction which we named {\sc
@@
 using statistical methods robust to outliers.
-
-\vspace{0.15cm}
 
@@
 
-\vspace{0.15cm}
-
 To address this, in our group we extended the first version of our algorithm {\sc ELUDE} to encompass modified peptides as well.
@@
 unmodified peptides  \cite{elude2}.
-
-\vspace{0.15cm}
 
@@
 
-\vspace{0.15cm}
-
 Palmblad et al. showed that matching theoretical and observed masses
@@
 
-\vspace{0.15cm}
-
 In our group we have also investigated the possibility to use predicted retention times to increase the throughput of proteomics experiments.
@@
 inference in shotgun proteomics \cite{mf}.
-
-\vspace{0.15cm}
 
@@
 discovery rate threshold of 1\% \cite{mf}.
-
-\vspace{0.15cm}
 
@@
 
-\vspace{0.15cm}
-
 Krokhin et al. illustrated yet another route to employ retention time
@@
 for identifying a set of proteins.
-
-\vspace{0.15cm}
 
@@
 
-\vspace{0.15cm}
-
 We tested this hypothesis by calculating and evaluating
@@
 applications.
-
-\vspace{0.15cm}
 
@@
 
-\vspace{0.15cm}
-
 As an example, in targeted proteomics peptide standards can be used to
@@
 
+\section{Discussion}
+
+Interpretation of mass spectrometry data is a question of
+understanding the properties that explain the behavior of peptides in
+an mass spectrometry experiment. The available technologies to infer
+peptides and proteins from mass spectrometry data involves predicting
+properties of peptides in a database, and compare those properties to
+the observed data. Such comparisons are for discovery proteomics done
+by so called search engines, that traditionally are limited to
+predicting predicting mass-to-charge rates of the precursor peptides,
+and their fragmentation patterns. However, an attractive way to
+increase the sensitivity of the inference procedure is to also predict
+properties such as their iso-electric point~\cite{branca2014hirief} or
+of particular interest for this review, chromatographic retention
+time\cite{cerqueira2010mude}.
+
+The principle of predictability also extends to targeted proteomics,
+in SRM analysis the peptides retention time and fragments are
+predicted from retention time, often derived from previous shotgun
+analysis. As targeted proteomics move into more comprehensive
+techniques, such as DIA/SWATH analysis, it becomes increasingly
+attractive to use techniques that do not rely on previous observations
+of the exact peptide to determine its properties, but instead to use
+de-novo predictions from amino acid sequences.
+
+% New technology to learn 
+% - ranking-based method
+% - deep learning
+
 \bibliographystyle{unsrt}

### 2014-11-23 retention.tex
Small fix in the Introduction

@@
 
-Although available in a variety of flavours, most of the widely spread MS-based workflows consists of the same main steps.
-First, the proteins in the mixture of interest are cleaved into peptides using a sequence-specific enzyme such as trypsin.
-This step is necessary because the mass spectrometers are more sensitive for peptides, and because spectra produced by full-length proteins are difficult to interpret.
-In the next step, the resulting peptides are separated according to their hydrophobicity on a liquid chromatography (LC) column.
-As they gradually elute from the column, the peptides are ionized via electrospray ionization and subjected to mass spectrometry analysis.
-The result of these measurements consists of a set of spectra, which are subsequently used to infer valuable information about the proteins present in the initial sample.
+Although available in a variety of flavours, most of the widely spread
+MS-based workflows consists of the same main steps. First, the
+proteins in the mixture of interest are cleaved into peptides using a
+sequence-specific enzyme such as trypsin. This step is necessary
+because the mass spectrometers are more sensitive for peptides, and
+because spectra produced by full-length proteins are difficult to
+interpret. In the next step, the resulting peptides are separated
+according to their hydrophobicity on a liquid chromatography (LC)
+column. As they gradually elute from the column, the peptides are
+ionized via electrospray ionization and subjected to mass spectrometry
+analysis. The result of these measurements consists of a set of
+spectra, which are subsequently used to infer valuable information
+about the proteins present in the initial sample.
 
@@
 reported~\cite{Lemeer2009}. To manage this formidable diversity,
-increasing efforts are made to extract as much information as possible
-from an experiment.
+increasing efforts are made to extract - besides the fragmentation
+spectra - additional information from each MS experiment.
 
-The main purpose of liquid chromatography is to reduce the number of
-peptides simultaneously injected into the mass spectrometer, and by
-this increase the proteome coverage achieved in an experiment.
-However, besides this, the separation process in itself provides
-useful information about the analytes. On the one hand, the instrument
-always records the time when a peptide elutes from the chromatography
-column, i.e. the retention time. On the other hand, the retention time
-can also be estimated by examining the sequence of the peptide. The
-availability of both predicted and observed retention times has proven
-to be a powerful feature, with several applications in mass
-spectrometry-based proteomics \cite{Palmblad2013}.
+An example of such additional information is represented by retention
+times, i.e. the time points when peptides elute from the liquid
+chromatography column.  The main purpose of liquid chromatography is
+to reduce the number of peptides simultaneously injected into the mass
+spectrometer, and by this increase the proteome coverage achieved in
+an experiment.  However, besides this, the separation process in
+itself provides useful information about the analytes. On the one
+hand, the instrument always records the time when a peptide elutes
+from the chromatography column, i.e. the retention time. On the other
+hand, the retention time can also be estimated by examining the
+sequence of the peptide. The availability of both predicted and
+observed retention times has proven to be a powerful feature, with
+several applications in mass spectrometry-based proteomics
+\cite{Palmblad2013}.
 

### 2014-11-23 retention.tex
Added some text about retention time standards

@@
 
-\cite{irt}
+These methods involve keeping a database of previously observed
+retention times for a set of peptides, and use these values as a basis
+for retention time predictions. In particular, in the past years the
+use of retention standards has become increasingly
+widespread \cite{olegstd, irt}. Such standards typically consist of a
+small number of peptides covering a wide span of hydrophobicities and
+with well studied behaviors under LC-MS/MS. These peptides are often
+analyzed together with the sample of interest, and by comparing their
+observed and expected behaviors one can obtain valuable information
+for a variety of applications.
+
+\vspace{0.15cm}
+
+As an example, in targeted proteomics, peptide standards can be used to
+ calibrate predicted retention times and to estimate the accuracy of
+ the prediction algorithms~\cite{Kiyonami01022011}. In addition,
+ recent studies have reported excellent results when using peptide
+ standards to adjust the time when certain transitions are scheduled
+ ``on-the-fly''~\cite{seb, irt}. Retention standards are also useful
+ for aligning data from multiple experiments~\cite{petritis2003}, to
+ transfer retention times between different chromatography
+ setups~\cite{seb}, or to keep track of the performance of LC-MS/MS
+ systems~\cite{qcal}.
 

### 2014-11-25 retention.tex
Added citation to Oleg's last article

@@
 
-These methods can be roughly divided into four classes: index-based
-methods, modeling-based techniques, machine learning approaches and
-look-up tables-based approaches. The main ideas underlying each
+These methods can be roughly divided into four classes: look-up
+tables-based approaches, index-based methods, modeling-based methods
+and machine learning techniques. The main ideas underlying each
 category are described below, while Table \ref{tab:rtmethods} lists a
@@
 These methods involve keeping a database of previously observed
-retention times for a set of peptides, and use these values as a basis
-for retention time predictions. In particular, in the past years the
-use of retention standards has become increasingly
+retention times for a set of peptides, and subsequently use these
+values as a basis for retention time predictions. In particular, in
+the past years the use of retention standards has become increasingly
 widespread \cite{olegstd, irt}. Such standards typically consist of a
@@
 % for a few other column
-Following these observations, Krokhin et al. developed
-SSRCalc \cite{Krokhin2004}, which is currently the most widely used
+Following these observations, Krokhin et al. developed SSRCalc
+\cite{Krokhin2004}, which is currently the most widely used
 index-based retention time predictor. In the initial version of the
@@
 nearest-neighbor effects of charged peptides and propensity to form
-helical structures \cite{Krokhin2006}. However, a limitation of the
-approach is that the algorithm was optimized for predicting retention
-times for certain chromatography setups. Thus, SSRCalc's prediction
-ability deteriorates for experimental conditions diverging from the
-setup on which the algorithm was optimized \cite{Spicer2007}.
+helical structures \cite{Krokhin2006}. Recently, it was showed that
+interactions involving helical structures are highly complex, and thus
+more sophisticated models are needed to account for the effects that
+such interactions have on peptide retention time \cite{ah}.
+
+A limitation of SSRCacls is that the algorithm was
+optimized for predicting retention times for certain chromatography
+setups. Thus, SSRCalc's prediction ability deteriorates for
+experimental conditions diverging from the setup on which the
+algorithm was optimized \cite{Spicer2007}.
 

### 2014-11-30 retention.tex
Small textual changes

@@
 
-Mass spectrometry-based methods are unquestionably the most popular
-techniques for studying the proteint content of biological samples.
+Mass spectrometry (MS) based methods are unquestionably the most popular
+techniques for studying the protein content of biological samples.
 The palette of protemic applications where such methods are used is
@@
 reported~\cite{Lemeer2009}. To manage this formidable diversity,
-increasing efforts are made to extract - besides the fragmentation
-spectra - additional information from each MS experiment.
+increasing efforts are made to extract besides the fragmentation
+spectra additional information from each MS experiment.
 
@@
 Following these observations, a variety of retention time prediction
-methods has been proposed. Most naturally one could
-
-These methods can be roughly divided into four classes: look-up
-tables-based approaches, index-based methods, modeling-based methods
-and machine learning techniques. The main ideas underlying each
-category are described below, while Table \ref{tab:rtmethods} lists a
-few of the most popular retention time prediction methods from each
-category.
+methods has been proposed. Most naturally one could divide these
+methods into four classes: look-up tables-based approaches,
+index-based methods, modeling-based methods and machine learning
+techniques. The main ideas underlying each category are described
+below, while Table \ref{tab:rtmethods} lists a few of the most popular
+retention time prediction methods from each category.
 
@@
 
-\section{\label{sec:irt}Peptide retention standards}
+%\section{\label{sec:irt}Peptide retention standards}
 
-\hyphenation{wide-spread}
+%\hyphenation{wide-spread}
 % peptide standards what they are, what they are used for 
-In the past years, the use of retention standards has become
-increasingly widespread in the context of mass spectrometry-based
-proteomics~\cite{olegstd, irt}. Such standards typically consist of a
-small number of peptides covering a wide span of hydrophobicities and
-with well studied behaviors under LC-MS/MS. These peptides are often
-analyzed together with the sample of interest, and by comparing their
-observed and expected behaviors one can obtain valuable information
-for a variety of applications.
+%In the past years, the use of retention standards has become
+%increasingly widespread in the context of mass spectrometry-based
+%proteomics~\cite{olegstd, irt}. Such standards typically consist of a
+%small number of peptides covering a wide span of hydrophobicities and
+%with well studied behaviors under LC-MS/MS. These peptides are often
+%analyzed together with the sample of interest, and by comparing their
+%observed and expected behaviors one can obtain valuable information
+%for a variety of applications.
 
-As an example, in targeted proteomics peptide standards can be used to
- calibrate predicted retention times and to estimate the accuracy of
- the prediction algorithms~\cite{Kiyonami01022011}. In addition,
- recent studies have reported excellent results when using peptide
- standards to adjust the time when certain transitions are scheduled
- ``on-the-fly''~\cite{seb, irt}. Retention standards are also useful
- for aligning data from multiple experiments~\cite{petritis2003}, to
- transfer retention times between different chromatography
- setups~\cite{seb}, or to keep track of the performance of LC-MS/MS
- systems~\cite{qcal}.
+%As an example, in targeted proteomics peptide standards can be used to
+% calibrate predicted retention times and to estimate the accuracy of
+% the prediction algorithms~\cite{Kiyonami01022011}. In addition,
+% recent studies have reported excellent results when using peptide
+% standards to adjust the time when certain transitions are scheduled
+% ``on-the-fly''~\cite{seb, irt}. Retention standards are also useful
+% for aligning data from multiple experiments~\cite{petritis2003}, to
+% transfer retention times between different chromatography
+% setups~\cite{seb}, or to keep track of the performance of LC-MS/MS
+% systems~\cite{qcal}.
 
@@
 understanding the properties that explain the behavior of peptides in
-an mass spectrometry experiment. The available technologies to infer
-peptides and proteins from mass spectrometry data involves predicting
-properties of peptides in a database, and compare those properties to
-the observed data. Such comparisons are for discovery proteomics done
-by so called search engines, that traditionally are limited to
+a mass spectrometry experiment. The available technologies to infer
+peptides and proteins from mass spectrometry data involve predicting
+properties of peptides from a database, and compare those properties to
+the observed data. Such comparisons are done for discovery proteomics
+by algorithms called search engines, that traditionally are limited to
 predicting predicting mass-to-charge rates of the precursor peptides,
@@
 
+% This paragraph is consfusing, should be rephrased
 The principle of predictability also extends to targeted proteomics,

### 2014-12-04 retention.tex
tired

@@
 peptides and proteins from mass spectrometry data involve predicting
-properties of peptides from a database, and compare those properties to
-the observed data. Such comparisons are done for discovery proteomics
-by algorithms called search engines, that traditionally are limited to
-predicting predicting mass-to-charge rates of the precursor peptides,
-and their fragmentation patterns. However, an attractive way to
-increase the sensitivity of the inference procedure is to also predict
-properties such as their iso-electric point~\cite{branca2014hirief} or
-of particular interest for this review, chromatographic retention
-time\cite{cerqueira2010mude}.
+properties of peptides from a database, and compare those properties
+to the observed data. Such comparisons are done for discovery
+proteomics by algorithms called search engines, that traditionally are
+limited to predicting predicting mass-to-charge rates of the precursor
+peptides, and their fragmentation patterns. However, an attractive way
+to increase the sensitivity of the inference procedure is to also
+predict properties such as their iso-electric
+point~\cite{branca2014hirief} or of particular interest for this
+review, chromatographic retention time\cite{cerqueira2010mude}.
 
-% This paragraph is consfusing, should be rephrased
-The principle of predictability also extends to targeted proteomics,
-in SRM analysis the peptides retention time and fragments are
-predicted from retention time, often derived from previous shotgun
-analysis. As targeted proteomics move into more comprehensive
-techniques, such as DIA/SWATH analysis, it becomes increasingly
-attractive to use techniques that do not rely on previous observations
-of the exact peptide to determine its properties, but instead to use
-de-novo predictions from amino acid sequences.
+Moreover, targeted proteomics experiments, such as Selected Reaction
+Monitoring (SRM) experiments, are dependent on the predictability of a
+peptides behavior. In such experiments the retention times and the
+suitable transitions are needed as input for the analysis. Frequently,
+such information is derived from previous shotgun analysis. However,
+as the capacity of targeted proteomics experiments expands by the
+introduction of the more comprehensive data independent acquisition
+techniques techniques (a.k.a. SWATH analysis), the pre-identification
+of peptide transitions and retention times threatens to become a
+limiting factor, as the analysis never can become more comprehensive
+than the previous shotgun analysis would be. In such a scenario it
+becomes increasingly attractive to predict retention time and
+transition using {\em de novo} predictions from the peptides amino
+acid sequences.  A system of identifications not dependent on previous
+peptide identifications from shotgun proteomics experiment would lead
+to a more direct experimental design and hopefully to more
+comprehensie coverage. Thus {\em de novo} retention time predictions
+have an important role to fill. 
+%However, to do so we need to make the technique more accurate.
 
-% New technology to learn 
-% - ranking-based method
-% - deep learning
+We can see that there are a number of interesting machine learning
+techniques that not yet have been used on the problem of predicting
+peptides retention time. Here we will bring up one such techniques,
+that we feel are particularly good matches, that we have not yet seen
+implemented for the problem. Pairwise Machine-learned ranking (MLR)
+(\url{http://dx.doi.org/10.1561%2F1500000016}) models can be trained
+to determine which of two examples that should be better ranked thatn
+the other. MLR is traditionally used for construction of ranking
+models for information retrieval systems. However, in the case of
+retention time prediction we can use such a predictor to predict the
+elusion order for sets of peptides, which can be useful in its own
+(coon; just in time), but can also be converted into absolute
+retention times, given that the distribution of their retention times
+are known.
+
+It is also worth noting that the current trend of processing huge data
+repositories, frequently referred to as bigdata, have put quite some
+intest in so called deep learning techniques such as belief networks
+using restricted multiple layered bolzman machines (RBM; cite
+http://bit.ly/1vMq6az) or deep neural networks. The idea is that the
+performance of predictors often are enhanced if one can structure the
+prediction into multiple layers. The drawback of such models are that
+they require large training seta. Given the very large sets of
+peptides with retention times that are assembled in various
+repositories, such as peptide atlas and pride, such methods could
+probably be implemented for retention time prediction as well.
 

### 2014-12-16 retention.tex
Added some references, and modified some sentences

--- 2014-11-30 retention.tex
@@
 
-Mass spectrometry (MS) based methods are unquestionably the most popular
-techniques for studying the protein content of biological samples.
-The palette of protemic applications where such methods are used is
-getting increasingly diversified, ranging from identifying and
-quantifying proteins in biological samples \cite{Geiger2012} to
-determining posttranslational modifications \cite{Huttlin2010} and
-elucidating protein interaction networks \cite{Gavin2011}.
+Mass spectrometry (MS) based methods are unquestionably the most
+popular techniques for analyzing the protein content of biological
+mixtures.  The ever-expanding palette of applications range from
+identifying and quantifying proteins in biological samples
+\cite{Geiger2012} to determining posttranslational modifications
+\cite{Huttlin2010} and elucidating protein interaction networks
+\cite{Gavin2011}.
 
-Although available in a variety of flavours, most of the widely spread
+Although available in a variety of flavors, most of the widely spread
 MS-based workflows consists of the same main steps. First, the
@@
 \subsection{Index-based methods}
+\label{sec:irt}
 \hyphenation{lear-ning}
@@
 BioLCCC & \url{http://theorchromo.ru/} & \cite{gorshkov2006} \\
-ANN & Link? & \cite{petritis2003 ,petritis2006improved} \\
+ANN & Link? & \cite{petritis2003,petritis2006improved} \\
 RTModel, RTPredict &  \url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTPredict.html}, 
@@
 
-A limitation of SSRCacls is that the algorithm was
-optimized for predicting retention times for certain chromatography
-setups. Thus, SSRCalc's prediction ability deteriorates for
-experimental conditions diverging from the setup on which the
-algorithm was optimized \cite{Spicer2007}.
+A limitation of SSRCalc is that the algorithm was optimized for
+predicting retention times for certain chromatography setups. Thus,
+SSRCalc's prediction ability deteriorates for experimental conditions
+diverging from the setup on which the algorithm was optimized
+\cite{Spicer2007}.
 
@@
 $\sim$5000 peptides originating from another organism. The algorithm
-reported good accuracies, especially taking into account that it only
+reported good accuracy, especially taking into account that it only
 used information about the amino acid composition of the peptides.
@@
 In a follow-up work, the method was extended by representing each
- peptide by 1052 instead of 20
- features \cite{petritis2006improved}. The new descriptors encoded the
- identity of each amino acid at each position in a sequence, as well
- as the length and the hydrophobic moment of the peptide. The new
- neural network was trained on more than 300,000 peptide sequences. To
- obtain such a large number of peptides with known retention times, a
- strategy to normalize elution times was developed. This allowed
- pooling of peptides identified in more than 12,000 shotgun
- experiments. The trained algorithm reported excellent performance for
- the 1303 peptides on which it was evaluated.
+peptide by 1052 instead of 20 features
+\cite{petritis2006improved}. The new descriptors encoded the identity
+of each amino acid at each position in a sequence, as well as the
+length and the hydrophobic moment of the peptide. The new neural
+network was trained on more than 300,000 peptide sequences. To obtain
+such a large number of peptides with known retention times, a strategy
+to normalize elution times was developed. This allowed pooling of
+peptides identified in more than 12,000 shotgun experiments. The
+trained algorithm reported excellent performance for the 1303 peptides
+on which it was evaluated.
 
@@
 chromatographic conditions, but yielded lower prediction
-accuracies. Subsequently, a more elaborate SVM-based method using a
+accuracy. Subsequently, a more elaborate SVM-based method using a
 customized kernel function was proposed \cite{rtpredict, rtpredictImproved}. This
@@
 Elude} \cite{elude1, elude2}. One of the main novelty
-points of {\sc Elude} wass the calculation of peptide features based on
+points of {\sc Elude} was the calculation of peptide features based on
 both a general hydrophobicity index, as well as a retention index
@@
 of the retention time prediction methods currently available is able
-to provide accuracies comparable to the small shifts observed when
+to provide accuracy comparable to the small shifts observed when
 running a sample twice through a chromatography column
@@
 
-To address this, in our group we extended the first version of our algorithm {\sc ELUDE} to encompass modified peptides as well.
-This was straightforward to achieve once we observed that the in-house trained  retention index contained nearly all the information incorporated in the more general hydrophobicity scale.
-This meant that we could remove all the features based on this index with minimal loss in accuracy.
-Further, we redefined all features describing a peptide in terms of the in-house retention index alone.
-Since this index was trained for each dataset at hand, the algorithm became fully trainable for any modified amino acid (Figure \ref{fig:elude}).
-We evaluated the new version of {\sc ELUDE} on peptides containing artifactual modifications introduced as part of an MS protocol called COmbined
-FRActional DIagonal Chromatography (COFRADIC) \cite{Gevaert2002}. Our results showed that  {\sc ELUDE} yielded equally accurate predictions for modified and
-unmodified peptides  \cite{elude2}.
+To address this, in our group we extended the first version of our
+algorithm {\sc ELUDE} to encompass modified peptides as well. This was
+straightforward to achieve once we observed that the in-house trained
+retention index contained nearly all the information incorporated in
+the more general hydrophobicity scale. This meant that we could remove
+all the features based on this index with minimal loss in
+accuracy. Further, we redefined all features describing a peptide in
+terms of the in-house retention index alone. Since this index was
+trained for each dataset at hand, the algorithm became fully trainable
+for any modified amino acid (Figure \ref{fig:elude}). We evaluated the
+new version of {\sc ELUDE} on peptides containing artifactual
+modifications introduced as part of an MS protocol called COmbined
+FRActional DIagonal Chromatography (COFRADIC) \cite{Gevaert2002}. Our
+results showed that {\sc ELUDE} yielded equally accurate predictions
+for modified and unmodified peptides \cite{elude2}.
 
@@
 peptides and proteins from mass spectrometry data involve predicting
-properties of peptides from a database, and compare those properties to
-the observed data. Such comparisons are done for discovery proteomics
-by algorithms called search engines, that traditionally are limited to
-predicting predicting mass-to-charge rates of the precursor peptides,
-and their fragmentation patterns. However, an attractive way to
-increase the sensitivity of the inference procedure is to also predict
-properties such as their iso-electric point~\cite{branca2014hirief} or
-of particular interest for this review, chromatographic retention
-time\cite{cerqueira2010mude}.
+properties of peptides from a database, and compare those properties
+to the observed data. Such comparisons are done for discovery
+proteomics by algorithms called search engines, that traditionally are
+limited to predicting predicting mass-to-charge rates of the precursor
+peptides, and their fragmentation patterns. However, an attractive way
+to increase the sensitivity of the inference procedure is to also
+predict properties such as their iso-electric
+point~\cite{branca2014hirief} or of particular interest for this
+review, chromatographic retention time\cite{cerqueira2010mude}.
 
-% This paragraph is consfusing, should be rephrased
-The principle of predictability also extends to targeted proteomics,
-in SRM analysis the peptides retention time and fragments are
-predicted from retention time, often derived from previous shotgun
-analysis. As targeted proteomics move into more comprehensive
-techniques, such as DIA/SWATH analysis, it becomes increasingly
-attractive to use techniques that do not rely on previous observations
-of the exact peptide to determine its properties, but instead to use
-de-novo predictions from amino acid sequences.
+Moreover, targeted proteomics experiments, such as Selected Reaction
+Monitoring (SRM) experiments, are dependent on the predictability of a
+peptides behavior. In such experiments the retention times and the
+suitable transitions are needed as input for the analysis. Frequently,
+such information is derived from previous shotgun analysis. As the
+capacity of targeted proteomics experiments expands by the
+introduction of the more comprehensive data independent acquisition
+techniques techniques\cite{Venable2004} (a.k.a. SWATH analysis), such
+pre-identification of peptide transitions and retention times
+threatens to become a limiting factor. Our current methods to identify
+peptides from DIA are only as comprehensive as the previous shotgun
+analysis would be. An attractive alternative is to predict retention
+time and transition using {\em de novo} predictions from the peptides
+amino acid sequences.  Such a system of identifications not dependent
+on previous peptide identifications from shotgun proteomics experiment
+would lead to a more direct experimental design and hopefully to more
+comprehensie coverage. Thus {\em de novo} retention time predictions
+have an important role to fill.
+%However, to do so we need to make the technique more accurate.
 
-% New technology to learn 
-% - ranking-based method
-% - deep learning
+We can see that there are a number of interesting machine learning
+techniques that not yet have been used on the problem of predicting
+peptides retention time. Here we will bring up one such techniques,
+that we feel are particularly good matches, that we have not yet seen
+implemented for the problem. Pairwise Machine-learned ranking (MLR)
+\cite{liu2009learning} models can be trained to determine which of two
+examples that should be better ranked thatn the other. MLR is
+traditionally used for construction of ranking models for information
+retrieval systems. However, in the case of retention time prediction
+we can use such a predictor to predict the elusion order for sets of
+peptides, which can be useful in its own \cite{bailey2012instant}, but
+can also be converted into absolute retention times, given that the
+distribution of their retention times are known.
+
+It is also worth noting that the current trend of processing huge data
+repositories, frequently referred to as bigdata, have put quite some
+intest in so called deep learning techniques such as belief networks
+using restricted multiple layered bolzman machines
+(RBM)\cite{salakhutdinov2009deep} or deep neural networks. This
+follows the observation that ideathe performance of predictors often
+are enhanced if one can structure the prediction into multiple
+layers. The drawback of such models are that they require large
+training sets. Given the very large sets of peptides with retention
+times that are assembled in various repositories, such as peptide
+atlas and pride, such methods could probably be implemented for
+retention time prediction as well.
+
+% Might be good to add something about that the relatively low
+% accuracy in our predictions, there should be some room for
+% improvements in the near future.
 
@@
 \end{document}
+
+%%  LocalWords:  hydrophobicity chromatography elute peptides physico
+%%  LocalWords:  electrospray spectrometry analytical proteomics RPLC
+%%  LocalWords:  posttranslational phosphorylation proteome analytes
+%%  LocalWords:  peptide elutes amino tryptic isoelectric SSRCacls
+%%  LocalWords:  workflows trypsin hydrophobicities chromatographic

### 2014-12-16 retention.tex
x

@@
 such predictions. Sections~\ref{sec:rtpred} and \ref{sec:rtpredm}
-describe the main approaches used to estimate retention time from
+describe the main approaches to retention time predictions from
 peptide sequences. Section~\ref{sec:app} discusses some of the most
 important applications of retention time predictions.
-Section~\ref{sec:irt} briefly introduces the use of peptide retention
-standards.
+% Section~\ref{sec:irt} briefly introduces the use of peptide retention
+% standards.
 
@@
 small number of peptides covering a wide span of hydrophobicities and
-with well studied behaviors under LC-MS/MS. These peptides are often
+favorable for detection by mass spectrometry. These peptides are often
 analyzed together with the sample of interest, and by comparing their
@@
 
-A limitation of SSRCalc is that the algorithm was optimized for
+A limitation of index-based methods are that they are optimized for
 predicting retention times for certain chromatography setups. Thus,
-SSRCalc's prediction ability deteriorates for experimental conditions
+their prediction ability deteriorates for experimental conditions
 diverging from the setup on which the algorithm was optimized
@@
 introduction of the more comprehensive data independent acquisition
-techniques techniques\cite{Venable2004} (a.k.a. SWATH analysis), such
+techniques techniques \cite{Venable2004} (a.k.a. SWATH analysis), such
 pre-identification of peptide transitions and retention times
@@
 using restricted multiple layered bolzman machines
-(RBM)\cite{salakhutdinov2009deep} or deep neural networks. This
+(RBM) \cite{salakhutdinov2009deep} or deep neural networks. This
 follows the observation that ideathe performance of predictors often

### 2015-01-06 retention.tex
x

@@
 \and
-Lukas K\"all\\
+Lukas K\"{a}ll\\
 Science for Life Laboratory,\\
@@
 Furthermore, a large proportion of proteins are subjected to
-posttranslational modifications. As an example, a few years ago more
-than 25,000 phosphorylation sites on human proteins were already
-reported~\cite{Lemeer2009}. To manage this formidable diversity,
-increasing efforts are made to extract besides the fragmentation
-spectra additional information from each MS experiment.
+posttranslational modifications~\cite{Lemeer2009}. To manage this
+formidable diversity, increasing efforts are made to extract besides
+the fragmentation spectra additional information from each MS
+experiment.
 
@@
 to reduce the number of peptides simultaneously injected into the mass
-spectrometer, and by this increase the proteome coverage achieved in
-an experiment.  However, besides this, the separation process in
-itself provides useful information about the analytes. On the one
-hand, the instrument always records the time when a peptide elutes
-from the chromatography column, i.e. the retention time. On the other
-hand, the retention time can also be estimated by examining the
-sequence of the peptide. The availability of both predicted and
-observed retention times has proven to be a powerful feature, with
-several applications in mass spectrometry-based proteomics
+spectrometer. This increases the probability of each individual
+peptide to be selected for fragmentation, which in its turn, increases
+the proteome coverage achieved in an experiment.  However, besides
+this, the separation process in itself provides useful information
+about the analytes. On the one hand, the instrument always records the
+time when a peptide elutes from the chromatography column, i.e. the
+retention time. On the other hand, the retention time can also be
+estimated by examining the sequence of the peptide. The availability
+of both predicted and observed retention times has proven to be a
+powerful feature, with several applications in MS-based proteomics
 \cite{Palmblad2013}.
@@
 
-In our group we developed a software package for retention time prediction which we named {\sc
-Elude} \cite{elude1, elude2}. One of the main novelty
-points of {\sc Elude} was the calculation of peptide features based on
-both a general hydrophobicity index, as well as a retention index
-trained for each dataset at hand. This ensured that the algorithm
-could be applied to different chromatography setups with minimal loss
-in performance. For cases when little training data was available,
-{\sc ELUDE} implemented an additional workflow based on a library of
-pre-trained retention models. This workflow required only a few tens
-of training peptides to select and calibrate the most suitable model
-from the library. Both the selection and calibration were performed
-using statistical methods robust to outliers.
+Moruz {\em et al.} developed a software package for retention time
+prediction which we named {\sc Elude} \cite{elude1, elude2}. One of
+the main novelty points of {\sc Elude} was the calculation of peptide
+features based on both a general hydrophobicity index, as well as a
+retention index trained for each dataset at hand. This ensured that
+the algorithm could be applied to different chromatography setups with
+minimal loss in performance. For cases when little training data was
+available, {\sc ELUDE} implemented an additional workflow based on a
+library of pre-trained retention models. This workflow required only a
+few tens of training peptides to select and calibrate the most
+suitable model from the library. Both the selection and calibration
+were performed using statistical methods robust to outliers.
 
@@
 
-Palmblad et al. showed that matching theoretical and observed masses
+Palmblad {\em et al.} showed that matching theoretical and observed masses
 and retention times provides a simple, yet powerful method to identify
@@
 
-In our group we have also investigated the possibility to use predicted retention times to increase the throughput of proteomics experiments.
-Our starting point was the observation that in typical shotgun proteomics experiments only a small fraction of the detected ions is
-fragmented, and thus used to identify
-peptides. The remaining unfragmented ions are usually disregarded in
-the subsequent data analyses. In the light of these observations, we investigated whether we could use
-information encapsulated by unfragmented ions to improve the protein
-inference in shotgun proteomics \cite{mf}.
+Moruz {\em et al.} also investigated the possibility to use predicted
+retention times to increase the throughput of proteomics
+experiments. Their starting point was the observation that in typical
+shotgun proteomics experiments only a small fraction of the detected
+ions is fragmented, and thus used to identify peptides. The remaining
+unfragmented ions are usually disregarded in the subsequent data
+analyses. In the light of these observations, they investigated
+whether we could use information encapsulated by unfragmented ions to
+improve the protein inference in shotgun proteomics \cite{mf}.
 
@@
 
+The current accuracy of the de novo retention time predictors that we
+have discussed here is still much lower than the methods based on look
+ups of previous registered retention time. The large availability of
+data leads one to think that there still is a possibility to increase
+the accuracy of such predictions by better technology. 
+% Shorten the next two paragraphs
+We can see that
+there are a number of interesting machine learning techniques that not
+yet have been used on the problem of predicting peptides retention
+time. Here we will bring up one such techniques, that we feel are
+particularly good matches, that we have not yet seen implemented for
+the problem. Pairwise Machine-learned ranking (MLR)
+\cite{liu2009learning} models can be trained to determine which of two
+examples that should be better ranked thatn the other. MLR is
+traditionally used for construction of ranking models for information
+retrieval systems. However, in the case of retention time prediction
+we can use such a predictor to predict the elusion order for sets of
+peptides, which can be useful in its own \cite{bailey2012instant}, but
+can also be converted into absolute retention times, given that the
+distribution of their retention times are known.
+
+It is also worth noting that the current trend of processing huge data
+repositories, frequently referred to as bigdata, have put quite some
+intest in so called deep learning techniques such as belief networks
+using restricted multiple layered bolzman machines
+(RBM) \cite{salakhutdinov2009deep} or deep neural networks. This
+follows the observation that ideathe performance of predictors often
+are enhanced if one can structure the prediction into multiple
+layers. The drawback of such models are that they require large
+training sets. Given the very large sets of peptides with retention
+times that are assembled in various repositories, such as peptide
+atlas and pride, such methods could probably be implemented for
+retention time prediction as well.
+
 Interpretation of mass spectrometry data is a question of
@@
 
-We can see that there are a number of interesting machine learning
-techniques that not yet have been used on the problem of predicting
-peptides retention time. Here we will bring up one such techniques,
-that we feel are particularly good matches, that we have not yet seen
-implemented for the problem. Pairwise Machine-learned ranking (MLR)
-\cite{liu2009learning} models can be trained to determine which of two
-examples that should be better ranked thatn the other. MLR is
-traditionally used for construction of ranking models for information
-retrieval systems. However, in the case of retention time prediction
-we can use such a predictor to predict the elusion order for sets of
-peptides, which can be useful in its own \cite{bailey2012instant}, but
-can also be converted into absolute retention times, given that the
-distribution of their retention times are known.
-
-It is also worth noting that the current trend of processing huge data
-repositories, frequently referred to as bigdata, have put quite some
-intest in so called deep learning techniques such as belief networks
-using restricted multiple layered bolzman machines
-(RBM) \cite{salakhutdinov2009deep} or deep neural networks. This
-follows the observation that ideathe performance of predictors often
-are enhanced if one can structure the prediction into multiple
-layers. The drawback of such models are that they require large
-training sets. Given the very large sets of peptides with retention
-times that are assembled in various repositories, such as peptide
-atlas and pride, such methods could probably be implemented for
-retention time prediction as well.
-
 % Might be good to add something about that the relatively low

### 2015-01-07 retention.tex
x

@@
 brought up a whole new set of analytical challenges, making the
-development of efficient algorithms to interpret this data a major
-area of proteomics research. A particular difficult challenge in this
-context is posed by the tremendous complexity of the data. Proteins
-display a wide range of physico-chemical properties and are often
-found at very different amounts in complex samples, spanning as much
-as 12 orders of magnitude in relative abundance~\cite{Angel2012}.
+development of efficient algorithms for data interpretation a major
+area within proteomics research. A particular difficult challenge in
+this context is posed by the tremendous complexity of the data.
+Proteins display a wide range of physico-chemical properties and are
+often found at very different amounts in complex samples, spanning as
+much as 12 orders of magnitude in relative abundance~\cite{Angel2012}.
 Furthermore, a large proportion of proteins are subjected to
 posttranslational modifications~\cite{Lemeer2009}. To manage this
-formidable diversity, increasing efforts are made to extract besides
-the fragmentation spectra additional information from each MS
-experiment.
+diversity, increasing efforts are made to extract besides the
+fragmentation spectra additional information from each MS experiment.
 
 An example of such additional information is represented by retention
-times, i.e. the time points when peptides elute from the liquid
+times, {\em i.e.} the time points when peptides elute from the liquid
 chromatography column.  The main purpose of liquid chromatography is
@@
 about the analytes. On the one hand, the instrument always records the
-time when a peptide elutes from the chromatography column, i.e. the
-retention time. On the other hand, the retention time can also be
+time when a peptide elutes from the chromatography column, {\em i.e.}
+the retention time. On the other hand, the retention time can also be
 estimated by examining the sequence of the peptide. The availability
@@
 \subsection{Machine learning-based methods}
+\label{sec:ML}
 
 In machine learning-based approaches a set of training peptides,
-i.e. peptide sequences with known retention times, are used to
+{\em i.e.} peptide sequences with known retention times, are used to
 customize a predefined mathematical model. This step is typically
@@
 
-Moruz {\em et al.} developed a software package for retention time
-prediction which we named {\sc Elude} \cite{elude1, elude2}. One of
-the main novelty points of {\sc Elude} was the calculation of peptide
-features based on both a general hydrophobicity index, as well as a
-retention index trained for each dataset at hand. This ensured that
-the algorithm could be applied to different chromatography setups with
+Moruz {\em et al.} \cite{elude1} developed a software package for
+retention time prediction which we named {\sc Elude}. One of the main
+novelty points of {\sc Elude} was the calculation of peptide features
+based on both a general hydrophobicity index, as well as a retention
+index trained for each dataset at hand. This ensured that the
+algorithm could be applied to different chromatography setups with
 minimal loss in performance. For cases when little training data was
-available, {\sc ELUDE} implemented an additional workflow based on a
+available, {\sc Elude} implemented an additional workflow based on a
 library of pre-trained retention models. This workflow required only a
@@
 
-To address this, in our group we extended the first version of our
-algorithm {\sc ELUDE} to encompass modified peptides as well. This was
-straightforward to achieve once we observed that the in-house trained
-retention index contained nearly all the information incorporated in
-the more general hydrophobicity scale. This meant that we could remove
-all the features based on this index with minimal loss in
-accuracy. Further, we redefined all features describing a peptide in
-terms of the in-house retention index alone. Since this index was
-trained for each dataset at hand, the algorithm became fully trainable
-for any modified amino acid (Figure \ref{fig:elude}). We evaluated the
-new version of {\sc ELUDE} on peptides containing artifactual
-modifications introduced as part of an MS protocol called COmbined
-FRActional DIagonal Chromatography (COFRADIC) \cite{Gevaert2002}. Our
-results showed that {\sc ELUDE} yielded equally accurate predictions
-for modified and unmodified peptides \cite{elude2}.
+To address this, in Moruz {\em et al.}\cite{elude2} extended the first
+version of their algorithm {\sc ELUDE} to encompass modified peptides
+as well. This was straightforward to achieve once the authors observed
+that the in-house trained retention index contained nearly all the
+information incorporated in the more general hydrophobicity scale.
+This meant that they could remove all the features based on this index
+with minimal loss in accuracy. Further, they redefined all features
+describing a peptide in terms of the in-house retention index alone.
+Since this index was trained for each dataset at hand, the algorithm
+became fully trainable for any modified amino acid (Figure
+\ref{fig:elude}). The new version of {\sc ELUDE} was evaluated on
+peptides containing artifactual modifications introduced as part of an
+MS protocol called COmbined FRActional DIagonal Chromatography
+(COFRADIC) \cite{Gevaert2002}. The results showed that {\sc ELUDE}
+yielded equally accurate predictions for modified and unmodified
+peptides \cite{elude2}.
 
@@
 
-The current accuracy of the de novo retention time predictors that we
-have discussed here is still much lower than the methods based on look
-ups of previous registered retention time. The large availability of
-data leads one to think that there still is a possibility to increase
-the accuracy of such predictions by better technology. 
-% Shorten the next two paragraphs
-We can see that
-there are a number of interesting machine learning techniques that not
-yet have been used on the problem of predicting peptides retention
-time. Here we will bring up one such techniques, that we feel are
-particularly good matches, that we have not yet seen implemented for
-the problem. Pairwise Machine-learned ranking (MLR)
-\cite{liu2009learning} models can be trained to determine which of two
-examples that should be better ranked thatn the other. MLR is
-traditionally used for construction of ranking models for information
-retrieval systems. However, in the case of retention time prediction
-we can use such a predictor to predict the elusion order for sets of
-peptides, which can be useful in its own \cite{bailey2012instant}, but
-can also be converted into absolute retention times, given that the
-distribution of their retention times are known.
+The accuracy of the machine learning-based methods described in
+section \ref{sec:ML} is still much lower than the methods based on
+look ups of previous registered retention time. However, the large
+availability of datasets leads one to think that there still is a
+possibility to increase the accuracy of such predictions by enhanced
+machinelearning technology.  We can see that there are a number of
+interesting techniques that not yet have been used on the problem of
+predicting peptides retention time. Here we will bring up one such
+techniques, that we feel are particularly good matches, that we have
+not yet seen implemented for the problem. Pairwise Machine-learned
+ranking (MLR) \cite{liu2009learning} models can be trained to
+determine which of two examples that should be better ranked thatn the
+other. MLR is traditionally used for construction of ranking models
+for information retrieval systems. However, in the case of retention
+time prediction we can use such a predictor to predict the elusion
+order for sets of peptides, which can be useful in its own
+\cite{bailey2012instant}, but can also be converted into absolute
+retention times, given that the distribution of their retention times
+are known.
 
@@
 intest in so called deep learning techniques such as belief networks
-using restricted multiple layered bolzman machines
-(RBM) \cite{salakhutdinov2009deep} or deep neural networks. This
-follows the observation that ideathe performance of predictors often
-are enhanced if one can structure the prediction into multiple
-layers. The drawback of such models are that they require large
-training sets. Given the very large sets of peptides with retention
-times that are assembled in various repositories, such as peptide
-atlas and pride, such methods could probably be implemented for
-retention time prediction as well.
+using restricted multiple layered bolzman machines (RBM)
+\cite{salakhutdinov2009deep} or deep neural networks. This follows the
+observation that ideathe performance of predictors often are enhanced
+if one can structure the prediction into multiple layers. The drawback
+of such models are that they require large training sets. Given the
+very large sets of peptides with retention times that are assembled in
+various repositories, such as Peptide Atlas \cite{desiere2006peptideatlas} and
+PRIDE \cite{martens2005pride}, such methods could probably be
+implemented for retention time prediction as well.
 
-Interpretation of mass spectrometry data is a question of
-understanding the properties that explain the behavior of peptides in
-a mass spectrometry experiment. The available technologies to infer
-peptides and proteins from mass spectrometry data involve predicting
-properties of peptides from a database, and compare those properties
-to the observed data. Such comparisons are done for discovery
-proteomics by algorithms called search engines, that traditionally are
-limited to predicting predicting mass-to-charge rates of the precursor
-peptides, and their fragmentation patterns. However, an attractive way
-to increase the sensitivity of the inference procedure is to also
-predict properties such as their iso-electric
-point~\cite{branca2014hirief} or of particular interest for this
-review, chromatographic retention time\cite{cerqueira2010mude}.
+The available technologies to infer peptides and proteins from mass
+spectrometry data involve predicting the the potential manifestations
+of the analyzed peptides, such as their fragmentation pattern, and
+comparing those predicted properties to the observed data. Such
+comparisons are regularly done for discovery proteomics by search
+engines. However, an attractive way to increase the sensitivity of the
+inference procedure is to also predict properties such as their
+iso-electric point~\cite{branca2014hirief} or of particular interest
+for this review, chromatographic retention
+time\cite{cerqueira2010mude}.
 
@@
 have an important role to fill.
-%However, to do so we need to make the technique more accurate.
 

### 2015-01-11 retention.tex
Added info about chromatography

@@
 time prediction, and describe a few of the successful applications of
-such predictions. Sections~\ref{sec:rtpred} and \ref{sec:rtpredm}
-describe the main approaches to retention time predictions from
-peptide sequences. Section~\ref{sec:app} discusses some of the most
-important applications of retention time predictions.
+such predictions. Section~\ref{sec:rplc} describes the basics of
+peptide separation using reversed-phase liquid chromatography, while
+Sections~\ref{sec:rtpred} and \ref{sec:rtpredm} summarize the main
+approaches to retention time predictions from peptide
+sequences. Section~\ref{sec:app} discusses some of the most important
+applications of retention time predictions.
 % Section~\ref{sec:irt} briefly introduces the use of peptide retention
 % standards.
+
+\section{\label{sec:rplc}Basics of peptide separation via reversed-phase liquid chromatography}
+% main idea about RPLC
+
+At the moment, Reversed-Phase Liquid Chromatography (RPLC) is by far
+the preferred method for separating peptides prior to mass
+spectrometry analysis \cite{Xie201278}. This is mainly because the
+method provides good resolution, gives reproducible results, and most
+of all, is compatible with electrospray ionization. In virtually all
+modern mass spectrometers the RPLC system can be easily connected to
+the ionization source, such that the peptides can be automatically
+transferred to mass spectrometry analysis as they elute from the
+chromatography column.
+
+In RPLC the peptides are separated according to their hydrophobicity
+in a thin tube called chromatographic column. This tube is uniformly
+packed with porous silica beads, such that there is as little
+unoccupied space as possible between the particles. Each of the beads
+contains hydrophobic chains attached. For peptide separation it is
+common to use long alkyl chains such as C8 or C18, although other
+options are also available. The packing of the column is commonly
+denoted as the stationary phase.
+
+Two types of solvents referred to as the mobile phase are used in
+RPLC. By convention, these solvents are denoted A and B. Solvent A
+refers to an aqueous buffer, typically composed of water and 0.1\%
+acid. Solvent B generally consists of a large percentage of organic
+solvent such as acetonitrile and 0.1\% acid. Any RPLC device includes
+a mechanism to pump and mix the two solvents at precise proportions.
+
+\begin{figure}[!h]
+\centering
+\includegraphics[trim=0.5cm 2.8cm 0.5cm 3.5cm, clip=true, width=0.8\textwidth]{img/rplc.pdf}
+\caption{\label{fig:rplc} {\bf Principles of liquid chromatography.}
+  The peptides mixed with mobile phase are injected on the column
+  (Panel A). As an increasing amount of solvent B is mixed in, the
+  peptides are released from the column in order of hydrophobicity
+  (Panels B and C). The signal at the end of the column can be
+  measured at each time point and displayed as a chromatogram (Panel
+  D). Image adapted from www.lamondlab.com.}
+\vspace{-7pt}
+\end{figure}
+
+During separation, the peptide mixture of interest is dissolved in
+solvent A and injected onto the column. As they move through the
+silica beads, peptides with different properties bind to the alkyl
+chains with varying strengths. To release them from the column, an
+increasing amount of solvent B is gradually mixed with buffer A and
+pumped through the column (Figure~\ref{fig:rplc}). Each peptide is
+released at a specific concentration of solvent B, and thus all the
+individual molecules representing a peptide species elute at about the
+same time.  Hence, if measuring the signal at the exit of the column,
+a peak would be observed for each distinct peptide species.
+
+The amount of time that a peptide needs to travel through the column
+is called retention time, shortly $t_R$. One often uses the term
+retention time to refer to the time corresponding to the apex of each
+peptide's elution peak (Figure~\ref{fig:rplc}, Panel D). Typically,
+the more hydrophobic the peptide, the longer the retention time.
+
+% say something about gradient and washing 
+Since the elution of the peptides is determined by the proportion of
+solvent B present at each time point, the scheme used to add solvent B
+is an essential aspect of any RPLC experiment. Currently, the most
+commonly used approach is based on so-called linear gradients, where
+the increase in solvent B follows a linear function. Most of the
+peptides elute before the percentage of solvent B reaches 40\%, and
+thus gradients often go from 2 to 40\% B. A high increase in
+percentage B at the end of the run ensures that all the remaining
+compounds are washed off from the column.
+
+A great advantage of RPLC is its remarkable versatility. By altering a
+variety of parameters such as column dimensions, the composition of
+stationary and mobile phase, temperature or gradient slope, the method
+can be adapted to separate a variety of compounds.
 

### 2015-01-11 retention.tex
Small textual changes in the Discussion

@@
 section \ref{sec:ML} is still much lower than the methods based on
-look ups of previous registered retention time. However, the large
-availability of datasets leads one to think that there still is a
-possibility to increase the accuracy of such predictions by enhanced
-machinelearning technology.  We can see that there are a number of
-interesting techniques that not yet have been used on the problem of
-predicting peptides retention time. Here we will bring up one such
-techniques, that we feel are particularly good matches, that we have
-not yet seen implemented for the problem. Pairwise Machine-learned
-ranking (MLR) \cite{liu2009learning} models can be trained to
-determine which of two examples that should be better ranked thatn the
-other. MLR is traditionally used for construction of ranking models
-for information retrieval systems. However, in the case of retention
-time prediction we can use such a predictor to predict the elusion
-order for sets of peptides, which can be useful in its own
-\cite{bailey2012instant}, but can also be converted into absolute
-retention times, given that the distribution of their retention times
-are known.
+look-ups of previously registered retention times. However, the large
+availability of datasets suggests that one possibility to increase the
+accuracy of retention time predictions could rely on the use of
+enhanced machine learning methods. Currently, there are a number of
+such techniques that have not yet been applied to the problem of
+predicting peptides' retention time. Here we will bring up one such
+technique called pairwise Machine-Learned Ranking (MLR)
+\cite{liu2009learning}, which we feel is particularly suitable for
+this problem. MLR models can be trained to determine which of two data
+points should be better ranked.
+
+MLR has traditionally been used for construction of ranking models for
+information retrieval systems. However, in the case of retention time
+prediction, we can use such a predictor to estimate the elution order
+for sets of peptides. While such information can be useful in its own
+\cite{bailey2012instant}, it can also be converted into absolute
+retention times, given that the distribution of the retention times is
+known.
 
 It is also worth noting that the current trend of processing huge data
-repositories, frequently referred to as bigdata, have put quite some
-intest in so called deep learning techniques such as belief networks
+repositories, frequently referred to as bigdata, have brought into
+attention so called deep learning techniques such as belief networks
 using restricted multiple layered bolzman machines (RBM)
 \cite{salakhutdinov2009deep} or deep neural networks. This follows the
-observation that ideathe performance of predictors often are enhanced
-if one can structure the prediction into multiple layers. The drawback
-of such models are that they require large training sets. Given the
-very large sets of peptides with retention times that are assembled in
-various repositories, such as Peptide Atlas \cite{desiere2006peptideatlas} and
-PRIDE \cite{martens2005pride}, such methods could probably be
-implemented for retention time prediction as well.
+observation that the performance of predictors is often enhanced if
+one can structure the prediction into multiple layers. The drawback of
+such models are that they require large training sets. Given the very
+large sets of peptides with retention times that are assembled in
+various repositories, such as Peptide Atlas
+\cite{desiere2006peptideatlas} and PRIDE \cite{martens2005pride}, such
+methods could probably be implemented for retention time prediction as
+well.
 
@@
 Moreover, targeted proteomics experiments, such as Selected Reaction
-Monitoring (SRM) experiments, are dependent on the predictability of a
-peptides behavior. In such experiments the retention times and the
-suitable transitions are needed as input for the analysis. Frequently,
+Monitoring (SRM), are dependent on the predictability of a
+peptide's behavior. In such experiments the retention times and the
+suitable transitions are needed as input to the analysis. Frequently,
 such information is derived from previous shotgun analysis. As the
@@
 introduction of the more comprehensive data independent acquisition
-techniques techniques \cite{Venable2004} (a.k.a. SWATH analysis), such
+techniques \cite{Venable2004} (a.k.a. SWATH analysis), such
 pre-identification of peptide transitions and retention times
@@
 amino acid sequences.  Such a system of identifications not dependent
-on previous peptide identifications from shotgun proteomics experiment
+on previous peptide identifications from shotgun proteomics experiments
 would lead to a more direct experimental design and hopefully to more
-comprehensie coverage. Thus {\em de novo} retention time predictions
+comprehensie coverages. Thus {\em de novo} retention time predictions
 have an important role to fill.

### 2015-01-13 retention.tex
Edits

@@
 retention time to refer to the time corresponding to the apex of each
-peptide's elution peak (Figure~\ref{fig:rplc}, Panel D). Typically,
-the more hydrophobic the peptide, the longer the retention time.
+peptide's elution peak (Figure~\ref{fig:rplc}, Panel D). This time
+point will be proportional to the peptide's ability to lower the
+surface potential between the beads and the solvent, a property that
+we normally refer to as the peptide's hydrophobicity.  Typically, the
+more hydrophobic the peptide, the longer the retention time.
 
@@
 BioLCCC & \url{http://theorchromo.ru/} & \cite{gorshkov2006} \\
-ANN & Link? & \cite{petritis2003,petritis2006improved} \\
+% ANN & Link? & \cite{petritis2003,petritis2006improved} \\
 RTModel, RTPredict &  \url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTPredict.html}, 
@@
 proteomics has gained popularity instead. Nevertheless, with great
-developments in instrumentation leading to sub-ppm mass accuracies,
+developments in instrumentation leading to sub-ppm mass accuracy,
 the possibility of using mass fingerprinting-like methods came back to
-attention~\cite{iamt1}. Our study showed how improved mass accuracies
+attention~\cite{iamt1}. Our study showed how improved mass accuracy
 and retention time predictions enable the use of unfragmented ions
@@
 
-We tested this hypothesis by calculating and evaluating
-two types of nonlinear gradients \cite{gradopt1}. Our results showed that the new
+We tested this hypothesis by calculating and evaluating two types of
+nonlinear gradients \cite{gradopt1}. Our results showed that the new
 gradients not only provided more uniform distributions of the
@@
 identifications. In addition, they facilitated the identification of
-different peptide species compared to the typical linear gradients.
-To allow the easy calculation of nonlinear gradients by any MS practitioner, we also implemented a user-friendly graphical user interface that can be downloaded free of cost \cite{gradopt2}.
-Seen from a general perspective, this showed that the gradient function is yet another parameter that researchers can alter
-in their experiments, opening up a whole new range of potential
-applications.
+different peptide species compared to the typical linear gradients. To
+allow the easy calculation of nonlinear gradients by any MS
+practitioner, we also implemented a user-friendly graphical user
+interface that can be downloaded free of cost \cite{gradopt2}.  Seen
+from a general perspective, this showed that the gradient function is
+yet another parameter that researchers can alter in their experiments,
+opening up a whole new range of potential applications.
 
@@
 attention so called deep learning techniques such as belief networks
-using restricted multiple layered bolzman machines (RBM)
+using restricted multiple layered Boltzmann machines (RBM)
 \cite{salakhutdinov2009deep} or deep neural networks. This follows the
@@
 would lead to a more direct experimental design and hopefully to more
-comprehensie coverages. Thus {\em de novo} retention time predictions
+comprehensive coverage. Thus {\em de novo} retention time predictions
 have an important role to fill.
@@
 %%  LocalWords:  posttranslational phosphorylation proteome analytes
-%%  LocalWords:  peptide elutes amino tryptic isoelectric SSRCacls
+%%  LocalWords:  peptide elutes amino tryptic isoelectric SSRCacls et
 %%  LocalWords:  workflows trypsin hydrophobicities chromatographic
+%%  LocalWords:  alkyl peptide's elution SSRCalc BioLCCC RTModel iRT
+%%  LocalWords:  RTPredict Krokhin al QSRR Klammer SVMs SVM Moruz pre
+%%  LocalWords:  dataset workflow outliers posttranslationally silico
+%%  LocalWords:  phosphorylated artifactual phosphopeptides COmbined
+%%  LocalWords:  FRActional DIagonal COFRADIC datasets Palmblad
+%%  LocalWords:  Strittmatter unfragmented

### 2015-01-25 retention.tex
Added overview image

@@
 \includegraphics[trim=0.5cm 0cm 2cm 1.5cm, clip=true, width=0.43\textwidth]{img/reproducibility.pdf}
-\caption{\label{fig:repr} {\bf Reproducibility of peptide retention time.} Two identical shotgun proteomics experiments were carried out.
-The retention times of the peptides confidently identified in both runs are displayed.}
+\caption{\label{fig:repr} {\bf Reproducibility of peptide retention
+    time.} Two identical shotgun proteomics experiments were carried
+  out. The retention times of the peptides confidently identified in
+  both runs are displayed.}
 %\end{figure}
-\vspace{-25pt}
+\vspace{-60pt}
 \end{wrapfigure}
@@
 index-based methods, modeling-based methods and machine learning
-techniques. The main ideas underlying each category are described
-below, while Table \ref{tab:rtmethods} lists a few of the most popular
-retention time prediction methods from each category.
+techniques. The main ideas underlying each category are illustrated in
+\ref{fig:methods} , while Table \ref{tab:rtmethods} lists a few of the
+most popular retention time prediction methods from each category.
+
+\begin{figure}[!h]
+\centering \includegraphics[clip=true,
+  width=0.8\textwidth]{img/methods.pdf}
+\caption{\label{fig:methods} {\bf Retention time prediction methods.}
+  Each figure illustrates a category of retention time prediction
+  methods. The details of each type of methods are described in detail
+  in Section \ref{sec:rtpred}}.
+\vspace{-7pt}
+\end{figure}
 

### 2015-01-26 retention.tex
Small textual changes

--- 2015-01-13 retention.tex
@@
 index-based methods, modeling-based methods and machine learning
-techniques. The main ideas underlying each category are described
-below, while Table \ref{tab:rtmethods} lists a few of the most popular
-retention time prediction methods from each category.
+techniques. The main ideas underlying each category are illustrated in
+Figure \ref{fig:methods} , while Table \ref{tab:rtmethods} lists a few
+of the most popular retention time prediction methods from each
+category.
+
+\begin{figure}[!h]
+\centering \includegraphics[clip=true,
+  width=0.8\textwidth]{img/methods.pdf}
+\caption{\label{fig:methods} {\bf Retention time prediction methods.}
+  Each figure illustrates a class of retention time prediction
+  methods. The details of the different methods are described in
+  detail in Section \ref{sec:rtpred}}.
+\vspace{-7pt}
+\end{figure}
 
@@
 Moruz {\em et al.} \cite{elude1} developed a software package for
-retention time prediction which we named {\sc Elude}. One of the main
-novelty points of {\sc Elude} was the calculation of peptide features
-based on both a general hydrophobicity index, as well as a retention
-index trained for each dataset at hand. This ensured that the
-algorithm could be applied to different chromatography setups with
-minimal loss in performance. For cases when little training data was
-available, {\sc Elude} implemented an additional workflow based on a
-library of pre-trained retention models. This workflow required only a
-few tens of training peptides to select and calibrate the most
-suitable model from the library. Both the selection and calibration
-were performed using statistical methods robust to outliers.
+retention time prediction named {\sc Elude}. One of the main novelty
+points of {\sc Elude} was the calculation of peptide features based on
+both a general hydrophobicity index, as well as a retention index
+trained for each dataset at hand. This ensured that the algorithm
+could be applied to different chromatography setups with minimal loss
+in performance. For cases when little training data was available,
+{\sc Elude} implemented an additional workflow based on a library of
+pre-trained retention models. This workflow required only a few tens
+of training peptides to select and calibrate the most suitable model
+from the library. Both the selection and calibration were performed
+using statistical methods robust to outliers.
 
@@
 of the retention time prediction methods currently available is able
-to provide accuracy comparable to the small shifts observed when
+to provide accuracies comparable to the small shifts observed when
 running a sample twice through a chromatography column
@@
 
-To address this, in Moruz {\em et al.}\cite{elude2} extended the first
-version of their algorithm {\sc ELUDE} to encompass modified peptides
-as well. This was straightforward to achieve once the authors observed
-that the in-house trained retention index contained nearly all the
-information incorporated in the more general hydrophobicity scale.
-This meant that they could remove all the features based on this index
-with minimal loss in accuracy. Further, they redefined all features
-describing a peptide in terms of the in-house retention index alone.
-Since this index was trained for each dataset at hand, the algorithm
-became fully trainable for any modified amino acid (Figure
+To address this, in Moruz {\em et al.}\cite{elude2} the version of
+their algorithm {\sc ELUDE} was extended to encompass modified
+peptides as well. This was straightforward to achieve once the authors
+observed that the in-house trained retention index contained nearly
+all the information incorporated in the more general hydrophobicity
+scale.  This meant that they could remove all the features based on
+this index with minimal loss in accuracy. Further, they redefined all
+features describing a peptide in terms of the in-house retention index
+alone.  Since this index was trained for each dataset at hand, the
+algorithm became fully trainable for any modified amino acid (Figure
 \ref{fig:elude}). The new version of {\sc ELUDE} was evaluated on
@@
 % how it works and main results
-More specifically, we matched observed masses and retention times of
+More specifically, they matched observed masses and retention times of
 unfragmented ions to predicted masses and retention times for the
-theoretical peptides from an {\it in silico} digest. We built a
+theoretical peptides from an {\it in silico} digest. They built a
 statistical framework to assign protein probabilities based on these
 matches, and combined the results with the ones obtained by the
-typical workflow based on fragmented ions. Following this approach, we
+typical workflow based on fragmented ions. Following this approach, they
 obtained up to 7\% more protein identifications at a fixed false
@@
 proteomics has gained popularity instead. Nevertheless, with great
-developments in instrumentation leading to sub-ppm mass accuracy,
-the possibility of using mass fingerprinting-like methods came back to
-attention~\cite{iamt1}. Our study showed how improved mass accuracy
-and retention time predictions enable the use of unfragmented ions
-even in the context of complex peptide mixtures.
+developments in instrumentation leading to sub-ppm mass accuracy, the
+possibility of using mass fingerprinting-like methods came back to
+attention~\cite{iamt1}. The mentioned study showed how improved mass
+accuracy and retention time predictions enable the use of unfragmented
+ions even in the context of complex peptide mixtures.
 
@@
 % Improve the design of SRM experiments 
-In addition, predicted retention times are useful investigating
-the elution properties of the peptides.  To illustrate this, we predicted the retention time of all the
-theoretical peptides from an {\it in silico} digest of the human
-proteome using  {\sc ELUDE}  \cite{elude1}. This exercise revealed a rather skewed
-distribution, with a majority of the peptides eluting in only a small
-portion of the chromatography run. This suggested that
-the linear gradient typically used in liquid chromatography are sub-optimal, and that one should be able to replace such gradients with nonlinear functions that produce even
-distributions of the peptides across runs.
+In addition, predicted retention times are useful investigating the
+elution properties of the peptides.  To illustrate this, in
+\cite{elude1} the retention time of all the theoretical peptides from
+an {\it in silico} digest of the human proteome were predicted using
+{\sc ELUDE}. This exercise revealed a rather skewed distribution, with
+a majority of the peptides eluting in only a small portion of the
+chromatography run. This suggested that the linear gradient typically
+used in liquid chromatography are sub-optimal, and that one should be
+able to replace such gradients with nonlinear functions that produce
+even distributions of the peptides across runs.
 
-We tested this hypothesis by calculating and evaluating two types of
-nonlinear gradients \cite{gradopt1}. Our results showed that the new
+This hypothesis was tested by calculating and evaluating two types of
+nonlinear gradients \cite{gradopt1}. The results showed that the new
 gradients not only provided more uniform distributions of the
@@
 allow the easy calculation of nonlinear gradients by any MS
-practitioner, we also implemented a user-friendly graphical user
-interface that can be downloaded free of cost \cite{gradopt2}.  Seen
-from a general perspective, this showed that the gradient function is
-yet another parameter that researchers can alter in their experiments,
-opening up a whole new range of potential applications.
+practitioner, a user-friendly graphical user interface was implemented 
+\cite{gradopt2}.  Seen from a general perspective, this showed that
+the gradient function is yet another parameter that researchers can
+alter in their experiments, opening up a whole new range of potential
+applications.
 
-Other examples of applications of retention time predictions include the optimization of  {\em de novo} targeted proteomic
-assays \cite{bertsch2010}, as well as decreasing the
-number of candidate peptides during database
-searching \cite{lobas2013} and accurate simulations of MS-based
-experiments \cite{bielow2011}.
+Other examples of applications of retention time predictions include
+the optimization of {\em de novo} targeted proteomic assays
+\cite{bertsch2010}, as well as decreasing the number of candidate
+peptides during database searching \cite{lobas2013} and accurate
+simulations of MS-based experiments \cite{bielow2011}.
 

### 2015-01-27 retention.tex
after spell-check

@@
 of the retention time prediction methods currently available is able
-to provide accuracies comparable to the small shifts observed when
-running a sample twice through a chromatography column
-(Figure \ref{fig:repr}). Recently, the lack of suitable
+to provide a prediction precision comparable to the small shifts
+observed when comparing retention time of peptides from two technical
+replicates (Figure \ref{fig:repr}). Recently, the lack of suitable
 representations for secondary structure elements was indicated as one

### 2015-01-27 retention.tex
small edits

@@
 proteins in the mixture of interest are cleaved into peptides using a
-sequence-specific enzyme such as trypsin. This step is necessary
-because the mass spectrometers are more sensitive for peptides, and
-because spectra produced by full-length proteins are difficult to
-interpret. In the next step, the resulting peptides are separated
-according to their hydrophobicity on a liquid chromatography (LC)
-column. As they gradually elute from the column, the peptides are
-ionized via electrospray ionization and subjected to mass spectrometry
-analysis. The result of these measurements consists of a set of
-spectra, which are subsequently used to infer valuable information
-about the proteins present in the initial sample.
+sequence-specific enzyme such as trypsin. In the next step, the
+resulting peptides are separated according to their hydrophobicity on
+a liquid chromatography (LC) column. As they gradually elute from the
+column, the peptides are ionized via electrospray ionization and
+subjected to mass spectrometry analysis. The result of these
+measurements consists of a set of spectra, which are subsequently used
+to infer valuable information about the proteins present in the
+initial sample.
 
-Today a majority of the mass spectrometry (MS) protocols in use are
+Today a majority of the mass spectrometry (MS) protocols in use have a
 high-throughput, generating extensive amounts of data. This has
@@
 
-In machine learning-based approaches a set of training peptides,
-{\em i.e.} peptide sequences with known retention times, are used to
-customize a predefined mathematical model. This step is typically
-called training or learning. Once trained, the mathematical model can
-be subsequently used to predict retention time for other peptides of
-interest.
+Machine learning-based approaches use a set of training peptides, {\em
+  i.e.} peptide sequences with known retention times, to estimate the
+parameters of a predefined mathematical model. This step is typically
+called fitting, training or learning. Once trained, the mathematical
+model can be subsequently used to predict retention time for other
+peptides of interest.
 
@@
 of the main causes of prediction errors \cite{Reimer2012}.  Most
-probably additional factors are involved, and further endeavors are
+probably additional factors are involved, and further studies are
 required to reach the desired level of accuracy.
@@
 peptides \cite{Kim2007}, or a couple of artifactual modifications
-commonly introduced during sample
-preparation \cite{Reimer2011}. Consequently, few retention time
-prediction methods have been extended and assessed for such
-peptides. A majority of the approaches currently available are
-modification-specific, many of them developed specifically for
-phosphopeptides \cite{Kawakami2005, perlova2010}.
+commonly introduced during sample preparation
+\cite{Reimer2011}. Consequently, few retention time prediction methods
+have been extended and assessed for such peptides. A majority of the
+approaches currently available are modification-specific, many of them
+developed specifically for phosphopeptides \cite{Kawakami2005,
+  perlova2010}.
 

### 2015-10-05 retention.tex
Answered first reviewer

--- 2015-01-26 retention.tex
@@
 time prediction, and describe a few of the successful applications of
-such predictions. Section~\ref{sec:rplc} describes the basics of
-peptide separation using reversed-phase liquid chromatography, while
-Sections~\ref{sec:rtpred} and \ref{sec:rtpredm} summarize the main
-approaches to retention time predictions from peptide
-sequences. Section~\ref{sec:app} discusses some of the most important
-applications of retention time predictions.
+such predictions. Sections~\ref{sec:rtpred} and \ref{sec:rtpredm}
+describe the main approaches to retention time predictions from
+peptide sequences. Section~\ref{sec:app} discusses some of the most
+important applications of retention time predictions.
 % Section~\ref{sec:irt} briefly introduces the use of peptide retention
 % standards.
-
-\section{\label{sec:rplc}Basics of peptide separation via reversed-phase liquid chromatography}
-% main idea about RPLC
-
-At the moment, Reversed-Phase Liquid Chromatography (RPLC) is by far
-the preferred method for separating peptides prior to mass
-spectrometry analysis \cite{Xie201278}. This is mainly because the
-method provides good resolution, gives reproducible results, and most
-of all, is compatible with electrospray ionization. In virtually all
-modern mass spectrometers the RPLC system can be easily connected to
-the ionization source, such that the peptides can be automatically
-transferred to mass spectrometry analysis as they elute from the
-chromatography column.
-
-In RPLC the peptides are separated according to their hydrophobicity
-in a thin tube called chromatographic column. This tube is uniformly
-packed with porous silica beads, such that there is as little
-unoccupied space as possible between the particles. Each of the beads
-contains hydrophobic chains attached. For peptide separation it is
-common to use long alkyl chains such as C8 or C18, although other
-options are also available. The packing of the column is commonly
-denoted as the stationary phase.
-
-Two types of solvents referred to as the mobile phase are used in
-RPLC. By convention, these solvents are denoted A and B. Solvent A
-refers to an aqueous buffer, typically composed of water and 0.1\%
-acid. Solvent B generally consists of a large percentage of organic
-solvent such as acetonitrile and 0.1\% acid. Any RPLC device includes
-a mechanism to pump and mix the two solvents at precise proportions.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[trim=0.5cm 2.8cm 0.5cm 3.5cm, clip=true, width=0.8\textwidth]{img/rplc.pdf}
-\caption{\label{fig:rplc} {\bf Principles of liquid chromatography.}
-  The peptides mixed with mobile phase are injected on the column
-  (Panel A). As an increasing amount of solvent B is mixed in, the
-  peptides are released from the column in order of hydrophobicity
-  (Panels B and C). The signal at the end of the column can be
-  measured at each time point and displayed as a chromatogram (Panel
-  D). Image adapted from www.lamondlab.com.}
-\vspace{-7pt}
-\end{figure}
-
-During separation, the peptide mixture of interest is dissolved in
-solvent A and injected onto the column. As they move through the
-silica beads, peptides with different properties bind to the alkyl
-chains with varying strengths. To release them from the column, an
-increasing amount of solvent B is gradually mixed with buffer A and
-pumped through the column (Figure~\ref{fig:rplc}). Each peptide is
-released at a specific concentration of solvent B, and thus all the
-individual molecules representing a peptide species elute at about the
-same time.  Hence, if measuring the signal at the exit of the column,
-a peak would be observed for each distinct peptide species.
-
-The amount of time that a peptide needs to travel through the column
-is called retention time, shortly $t_R$. One often uses the term
-retention time to refer to the time corresponding to the apex of each
-peptide's elution peak (Figure~\ref{fig:rplc}, Panel D). This time
-point will be proportional to the peptide's ability to lower the
-surface potential between the beads and the solvent, a property that
-we normally refer to as the peptide's hydrophobicity.  Typically, the
-more hydrophobic the peptide, the longer the retention time.
-
-% say something about gradient and washing 
-Since the elution of the peptides is determined by the proportion of
-solvent B present at each time point, the scheme used to add solvent B
-is an essential aspect of any RPLC experiment. Currently, the most
-commonly used approach is based on so-called linear gradients, where
-the increase in solvent B follows a linear function. Most of the
-peptides elute before the percentage of solvent B reaches 40\%, and
-thus gradients often go from 2 to 40\% B. A high increase in
-percentage B at the end of the run ensures that all the remaining
-compounds are washed off from the column.
-
-A great advantage of RPLC is its remarkable versatility. By altering a
-variety of parameters such as column dimensions, the composition of
-stationary and mobile phase, temperature or gradient slope, the method
-can be adapted to separate a variety of compounds.
 
@@
 virtually all the information needed to predict its retention time.
+However, in practice this is not as straight forward as it might seem, as it for instance is difficult to predict a peptide's three dimensional structure from its amino acid sequence.
 
-\begin{wrapfigure}{r}{.45\textwidth}
-\vspace{-10pt}
-%\begin{figure}[htbp]
+%\begin{wrapfigure}{r}{.45\textwidth}
+%\vspace{-10pt}
+\begin{figure}[htbp]
 \centering
 \includegraphics[trim=0.5cm 0cm 2cm 1.5cm, clip=true, width=0.43\textwidth]{img/reproducibility.pdf}
-\caption{\label{fig:repr} {\bf Reproducibility of peptide retention
-    time.} Two identical shotgun proteomics experiments were carried
-  out. The retention times of the peptides confidently identified in
-  both runs are displayed.}
-%\end{figure}
-\vspace{-60pt}
-\end{wrapfigure}
+\caption{\label{fig:repr} {\bf Reproducibility of peptide retention time.} Two identical shotgun proteomics experiments were carried out.
+The retention times of the peptides confidently identified in both runs are displayed.}
+\end{figure}
+%\vspace{-25pt}
+%\end{wrapfigure}
 
@@
 index-based methods, modeling-based methods and machine learning
-techniques. The main ideas underlying each category are illustrated in
-Figure \ref{fig:methods} , while Table \ref{tab:rtmethods} lists a few
-of the most popular retention time prediction methods from each
-category.
-
-\begin{figure}[!h]
-\centering \includegraphics[clip=true,
-  width=0.8\textwidth]{img/methods.pdf}
-\caption{\label{fig:methods} {\bf Retention time prediction methods.}
-  Each figure illustrates a class of retention time prediction
-  methods. The details of the different methods are described in
-  detail in Section \ref{sec:rtpred}}.
-\vspace{-7pt}
-\end{figure}
+techniques. The main ideas underlying each category are described
+below, while Table \ref{tab:rtmethods} lists a few of the most popular
+retention time prediction methods from each category.
 
@@
 BioLCCC & \url{http://theorchromo.ru/} & \cite{gorshkov2006} \\
-% ANN & Link? & \cite{petritis2003,petritis2006improved} \\
+ANN & Link? & \cite{petritis2003,petritis2006improved} \\
 RTModel, RTPredict &  \url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTPredict.html}, 
@@
 
-Elude & \url{http://elude.sbc.su.se/} & \cite{elude1, elude2} \\
+{\sc Elude} & \url{http://elude.sbc.su.se/} & \cite{elude1, elude2} \\
 iRT-calculator & http://www.biognosys.ch/products/irt-calculator.html
@@
 % for a few other column
-Following these observations, Krokhin et al. developed SSRCalc
+Following these observations, Krokhin {\em et al.} developed SSRCalc
 \cite{Krokhin2004}, which is currently the most widely used
@@
 setups is commonly used across research laboratories. To address this,
-Klammer et al. proposed a predictor based on support vector machines
+Klammer {\em et al.} proposed a predictor based on support vector machines
 (SVMs) \cite{klammer2007improving}. Their method required fewer
@@
 Moruz {\em et al.} \cite{elude1} developed a software package for
-retention time prediction named {\sc Elude}. One of the main novelty
-points of {\sc Elude} was the calculation of peptide features based on
-both a general hydrophobicity index, as well as a retention index
-trained for each dataset at hand. This ensured that the algorithm
-could be applied to different chromatography setups with minimal loss
-in performance. For cases when little training data was available,
-{\sc Elude} implemented an additional workflow based on a library of
-pre-trained retention models. This workflow required only a few tens
-of training peptides to select and calibrate the most suitable model
-from the library. Both the selection and calibration were performed
-using statistical methods robust to outliers.
+retention time prediction which we named {\sc Elude}. One of the main
+novelty points of {\sc Elude} was the calculation of peptide features
+based on both a general hydrophobicity index, as well as a retention
+index trained for each dataset at hand. This ensured that the
+algorithm could be applied to different chromatography setups with
+minimal loss in performance. For cases when little training data was
+available, {\sc Elude} implemented an additional workflow based on a
+library of pre-trained retention models. This workflow required only a
+few tens of training peptides to select and calibrate the most
+suitable model from the library. Both the selection and calibration
+were performed using statistical methods robust to outliers.
 
@@
 
-To address this, in Moruz {\em et al.}\cite{elude2} the version of
-their algorithm {\sc ELUDE} was extended to encompass modified
+To address this, in Moruz {\em et al.} \cite{elude2} extended the
+first version of their algorithm {\sc ELUDE} to encompass modified
 peptides as well. This was straightforward to achieve once the authors
-observed that the in-house trained retention index contained nearly
-all the information incorporated in the more general hydrophobicity
-scale.  This meant that they could remove all the features based on
-this index with minimal loss in accuracy. Further, they redefined all
-features describing a peptide in terms of the in-house retention index
-alone.  Since this index was trained for each dataset at hand, the
-algorithm became fully trainable for any modified amino acid (Figure
-\ref{fig:elude}). The new version of {\sc ELUDE} was evaluated on
-peptides containing artifactual modifications introduced as part of an
-MS protocol called COmbined FRActional DIagonal Chromatography
-(COFRADIC) \cite{Gevaert2002}. The results showed that {\sc ELUDE}
-yielded equally accurate predictions for modified and unmodified
-peptides \cite{elude2}.
+observed that many of their predictive features of their method were
+all functions of the peptides amino acid sequence and a retention
+index. As such retention indices are estimated for each dataset at
+hand, at the beginning of the training procedure, they can trivially
+be extended beyond the standard amino acids. When the authors used but
+the features based on their retention time index, they obtained a
+negligible loss in accuracy compared to using all their original
+features for unmodified peptides. Since this index was trained for
+each dataset at hand, the algorithm became fully trainable for any
+modified amino acid (Figure \ref{fig:elude}). The new version of {\sc
+ELUDE} was evaluated on peptides containing artifactual modifications
+introduced as part of an MS protocol called COmbined FRActional
+DIagonal Chromatography (COFRADIC) \cite{Gevaert2002}. The results
+showed that {\sc ELUDE} yielded equally accurate predictions for
+modified and unmodified peptides \cite{elude2}.
 
@@
 peptides in simple biological
-mixtures \cite{palmblad2002prediction}. Strittmatter et al. used
+mixtures \cite{palmblad2002prediction}. Strittmatter {\em et al.} used
 retention time as a parameter in a discriminant function that
@@
 reported $\sim$9\% more confident peptide identifications in a typical
-shotgun experiment. Along the same lines, Klammer et al. demonstrated
+shotgun experiment. Along the same lines, Klammer {\em et al.} demonstrated
 that for specific experimental conditions, just filtering out peptides
@@
 % how it works and main results
-More specifically, they matched observed masses and retention times of
+More specifically, we matched observed masses and retention times of
 unfragmented ions to predicted masses and retention times for the
-theoretical peptides from an {\it in silico} digest. They built a
+theoretical peptides from an {\it in silico} digest. We built a
 statistical framework to assign protein probabilities based on these
 matches, and combined the results with the ones obtained by the
-typical workflow based on fragmented ions. Following this approach, they
+typical workflow based on fragmented ions. Following this approach, we
 obtained up to 7\% more protein identifications at a fixed false
@@
 proteomics has gained popularity instead. Nevertheless, with great
-developments in instrumentation leading to sub-ppm mass accuracy, the
-possibility of using mass fingerprinting-like methods came back to
-attention~\cite{iamt1}. The mentioned study showed how improved mass
-accuracy and retention time predictions enable the use of unfragmented
-ions even in the context of complex peptide mixtures.
+developments in instrumentation leading to sub-ppm mass accuracies,
+the possibility of using mass fingerprinting-like methods came back to
+attention~\cite{iamt1}. Our study showed how improved mass accuracies
+and retention time predictions enable the use of unfragmented ions
+even in the context of complex peptide mixtures.
 
-Krokhin et al. illustrated yet another route to employ retention time
+Krokhin {\em et al.} illustrated yet another route to employ retention time
 predictions \cite{krokhin200612, McQueen2012}. They showed that
@@
 % Improve the design of SRM experiments 
-In addition, predicted retention times are useful investigating the
-elution properties of the peptides.  To illustrate this, in
-\cite{elude1} the retention time of all the theoretical peptides from
-an {\it in silico} digest of the human proteome were predicted using
-{\sc ELUDE}. This exercise revealed a rather skewed distribution, with
-a majority of the peptides eluting in only a small portion of the
-chromatography run. This suggested that the linear gradient typically
-used in liquid chromatography are sub-optimal, and that one should be
-able to replace such gradients with nonlinear functions that produce
-even distributions of the peptides across runs.
+In addition, predicted retention times are useful investigating
+the elution properties of the peptides.  To illustrate this, we predicted the retention time of all the
+theoretical peptides from an {\it in silico} digest of the human
+proteome using  {\sc Elude}  \cite{elude1}. This exercise revealed a rather skewed
+distribution, with a majority of the peptides eluting in only a small
+portion of the chromatography run. This suggested that
+the linear gradient typically used in liquid chromatography are sub-optimal, and that one should be able to replace such gradients with nonlinear functions that produce even
+distributions of the peptides across runs.
 
-This hypothesis was tested by calculating and evaluating two types of
-nonlinear gradients \cite{gradopt1}. The results showed that the new
+We tested this hypothesis by calculating and evaluating
+two types of nonlinear gradients \cite{gradopt1}. Our results showed that the new
 gradients not only provided more uniform distributions of the
@@
 identifications. In addition, they facilitated the identification of
-different peptide species compared to the typical linear gradients. To
-allow the easy calculation of nonlinear gradients by any MS
-practitioner, a user-friendly graphical user interface was implemented 
-\cite{gradopt2}.  Seen from a general perspective, this showed that
-the gradient function is yet another parameter that researchers can
-alter in their experiments, opening up a whole new range of potential
+different peptide species compared to the typical linear gradients.
+To allow the easy calculation of nonlinear gradients by any MS practitioner, we also implemented a user-friendly graphical user interface that can be downloaded free of cost \cite{gradopt2}.
+Seen from a general perspective, this showed that the gradient function is yet another parameter that researchers can alter
+in their experiments, opening up a whole new range of potential
 applications.
 
-Other examples of applications of retention time predictions include
-the optimization of {\em de novo} targeted proteomic assays
-\cite{bertsch2010}, as well as decreasing the number of candidate
-peptides during database searching \cite{lobas2013} and accurate
-simulations of MS-based experiments \cite{bielow2011}.
+Other examples of applications of retention time predictions include the optimization of  {\em de novo} targeted proteomic
+assays \cite{bertsch2010}, as well as decreasing the
+number of candidate peptides during database
+searching \cite{lobas2013} and accurate simulations of MS-based
+experiments \cite{bielow2011}.
 
@@
 section \ref{sec:ML} is still much lower than the methods based on
-look-ups of previously registered retention times. However, the large
-availability of datasets suggests that one possibility to increase the
-accuracy of retention time predictions could rely on the use of
-enhanced machine learning methods. Currently, there are a number of
-such techniques that have not yet been applied to the problem of
-predicting peptides' retention time. Here we will bring up one such
-technique called pairwise Machine-Learned Ranking (MLR)
-\cite{liu2009learning}, which we feel is particularly suitable for
-this problem. MLR models can be trained to determine which of two data
-points should be better ranked.
-
-MLR has traditionally been used for construction of ranking models for
-information retrieval systems. However, in the case of retention time
-prediction, we can use such a predictor to estimate the elution order
-for sets of peptides. While such information can be useful in its own
-\cite{bailey2012instant}, it can also be converted into absolute
-retention times, given that the distribution of the retention times is
-known.
+look ups of previous registered retention time. However, the large
+availability of datasets leads one to think that there still is a
+possibility to increase the accuracy of such predictions by enhanced
+machinelearning technology.  We can see that there are a number of
+interesting techniques that not yet have been used on the problem of
+predicting peptides retention time. Here we will bring up one such
+techniques, that we feel are particularly good matches, that we have
+not yet seen implemented for the problem. Pairwise Machine-learned
+ranking (MLR) \cite{liu2009learning} models can be trained to
+determine which of two examples that should be better ranked thatn the
+other. MLR is traditionally used for construction of ranking models
+for information retrieval systems. However, in the case of retention
+time prediction we can use such a predictor to predict the elusion
+order for sets of peptides, which can be useful in its own
+\cite{bailey2012instant}, but can also be converted into absolute
+retention times, given that the distribution of their retention times
+are known.
 
 It is also worth noting that the current trend of processing huge data
-repositories, frequently referred to as bigdata, have brought into
-attention so called deep learning techniques such as belief networks
-using restricted multiple layered Boltzmann machines (RBM)
+repositories, frequently referred to as big data, have put quite some
+intest in so called deep learning techniques such as belief networks
+using restricted multiple layered bolzman machines (RBM)
 \cite{salakhutdinov2009deep} or deep neural networks. This follows the
-observation that the performance of predictors is often enhanced if
-one can structure the prediction into multiple layers. The drawback of
-such models are that they require large training sets. Given the very
-large sets of peptides with retention times that are assembled in
-various repositories, such as Peptide Atlas
-\cite{desiere2006peptideatlas} and PRIDE \cite{martens2005pride}, such
-methods could probably be implemented for retention time prediction as
-well.
+observation that ideathe performance of predictors often are enhanced
+if one can structure the prediction into multiple layers. The drawback
+of such models are that they require large training sets. Given the
+very large sets of peptides with retention times that are assembled in
+various repositories, such as Peptide Atlas \cite{desiere2006peptideatlas} and
+PRIDE \cite{martens2005pride}, such methods could probably be
+implemented for retention time prediction as well.
 
@@
 for this review, chromatographic retention
-time\cite{cerqueira2010mude}.
+time \cite{cerqueira2010mude}.
 
 Moreover, targeted proteomics experiments, such as Selected Reaction
-Monitoring (SRM), are dependent on the predictability of a
-peptide's behavior. In such experiments the retention times and the
-suitable transitions are needed as input to the analysis. Frequently,
+Monitoring (SRM) experiments, are dependent on the predictability of a
+peptides behavior. In such experiments the retention times and the
+suitable transitions are needed as input for the analysis. Frequently,
 such information is derived from previous shotgun analysis. As the
@@
 introduction of the more comprehensive data independent acquisition
-techniques \cite{Venable2004} (a.k.a. SWATH analysis), such
+techniques techniques \cite{Venable2004} (a.k.a. SWATH analysis), such
 pre-identification of peptide transitions and retention times
 threatens to become a limiting factor. Our current methods to identify
-peptides from DIA are only as comprehensive as the previous shotgun
+peptides from Data Independent Acquisition are only as comprehensive as the previous shotgun
 analysis would be. An attractive alternative is to predict retention
@@
 amino acid sequences.  Such a system of identifications not dependent
-on previous peptide identifications from shotgun proteomics experiments
+on previous peptide identifications from shotgun proteomics experiment
 would lead to a more direct experimental design and hopefully to more
-comprehensive coverage. Thus {\em de novo} retention time predictions
+comprehensie coverage. Thus {\em de novo} retention time predictions
 have an important role to fill.
@@
 %%  LocalWords:  posttranslational phosphorylation proteome analytes
-%%  LocalWords:  peptide elutes amino tryptic isoelectric SSRCacls et
+%%  LocalWords:  peptide elutes amino tryptic isoelectric SSRCacls
 %%  LocalWords:  workflows trypsin hydrophobicities chromatographic
-%%  LocalWords:  alkyl peptide's elution SSRCalc BioLCCC RTModel iRT
-%%  LocalWords:  RTPredict Krokhin al QSRR Klammer SVMs SVM Moruz pre
-%%  LocalWords:  dataset workflow outliers posttranslationally silico
-%%  LocalWords:  phosphorylated artifactual phosphopeptides COmbined
-%%  LocalWords:  FRActional DIagonal COFRADIC datasets Palmblad
-%%  LocalWords:  Strittmatter unfragmented

### 2015-10-05 response.tex
Answered first reviewer

\documentclass[letterpaper]{article}
\usepackage{url}
\usepackage[margin=1in]{geometry}
\newcommand{\breview}{\begin{quotation}\begin{bf}\noindent}
\newcommand{\ereview}{\end{bf}\end{quotation}}

\begin{document}

\section*{Reviewer \#1}

\breview

The authors give a good overview of the existing methods to predict the retention time.  However, the description of the methods is sometimes too brief, and slightly biased towards their own developed methods. Extending the descriptions of the methods that are not developed by the authors should be considered. The applications and discussion section are an added value to the paper.

\ereview

We would like to thank the reviewer for the positive feedback. We have revised the sections were the other methods contribution is discussed

\breview
Additional comments:
\begin{itemize}
\item Legend of Figure 2 overlaps with text on page 3
\item Table 1 is not fully visible; additionally in the table “Elude” is written, while in the text “ELUDE” is used
\end{itemize}
\ereview
We have corrected this.
\breview
\begin{itemize}

\item The authors write on page 5 and 6: ``The algorithm reported good
accuracy, especially taking into account that it only used information
about the amino acid composition of the peptides.'' This statement
seems to be contrasting with the following comment found on page 3:
``As a consequence, given a chromatography setup, the sequence of a
peptide should provide virtually all the information needed to predict
its retention time''. To the latter statement several other
contrasting statements can be found on the same page.
\end{itemize}
\ereview
We added a qualifying paragraph to the later statement.
``As a consequence, given a chromatography setup, the sequence of a
peptide should provide virtually all the information needed to predict
its retention time. However, in practice this is not as straight
forward as it might seem, as it for instance is difficult to predict a
peptide's three dimensional structure from its amino acid sequence.''
\breview
\begin{itemize}
\item The paragraph in Section 4 on the modified ELUDE is not completely clear. An extended discussion would be useful
\end{itemize}
\ereview

We revised this section to better explain how we incorporated
retention time predictions of modified peptides.

\breview
Typos:
\begin{itemize}
\item Page 6: Klammer et al. $\to$ Klammer {\em et al.}
\item Page 6: Moruz et al.[22] $\to$ Moruz {\em et al.} [22]
\item Page 7: Strittmatter et al. $\to$ Strittmatter {\em et al.}
\item Page 7: Klammer et al. $\to$ Klammer {\em et al.}
\item Page 8: Krokhin et al. $\to$ Krokhin {\em et al.}
\item Page 8: bigdata $\to$ big data
\item Page 9: retention time[57] $\to$ retention time [57].
\item Page 9: DIA $\to$ data independent acquisition
\end{itemize}

\ereview

Thank you for highlighting these typos. We have now corrected them.

\section*{Reviewer \#2}

\breview
\ereview

\end{document}

### 2015-10-05 retention.tex

--- 2015-01-27 retention.tex
@@
 virtually all the information needed to predict its retention time.
+However, in practice this is not as straight forward as it might seem, as it for instance is difficult to predict a peptide's three dimensional structure from its amino acid sequence.
 
-\begin{wrapfigure}{r}{.45\textwidth}
-\vspace{-10pt}
-%\begin{figure}[htbp]
+%\begin{wrapfigure}{r}{.45\textwidth}
+%\vspace{-10pt}
+\begin{figure}[htbp]
 \centering
@@
   both runs are displayed.}
-%\end{figure}
-\vspace{-60pt}
-\end{wrapfigure}
+\end{figure}
+%\vspace{-25pt}
+%\end{wrapfigure}
 
@@
 
-Krokhin et al. illustrated yet another route to employ retention time
+Krokhin {\em et al.} illustrated yet another route to employ retention time
 predictions \cite{krokhin200612, McQueen2012}. They showed that
@@
 It is also worth noting that the current trend of processing huge data
-repositories, frequently referred to as bigdata, have brought into
+repositories, frequently referred to as big data, have brought into
 attention so called deep learning techniques such as belief networks
@@
 for this review, chromatographic retention
-time\cite{cerqueira2010mude}.
+time \cite{cerqueira2010mude}.
 
@@
 threatens to become a limiting factor. Our current methods to identify
-peptides from DIA are only as comprehensive as the previous shotgun
+peptides from Data Independent Acquisition are only as comprehensive as the previous shotgun
 analysis would be. An attractive alternative is to predict retention

### 2015-10-06 response.tex
have to leave

@@
 \item Legend of Figure 2 overlaps with text on page 3
-\item Table 1 is not fully visible; additionally in the table “Elude” is written, while in the text “ELUDE” is used
+\item Table 1 is not fully visible; additionally in the table ``Elude'' is written, while in the text ``ELUDE'' is used
 \end{itemize}
@@
 \breview
+
+The reviewer feels the paper’s target audience isn’t clear. As such,
+the overall sensation is that the paper touches several subjects
+lightly but fails in going in detail into any of them. As it is right
+now, this paper is not practical enough for a mass spectrometrist, and
+not technical enough for a machine learning scientist. The reviewer
+feels it is necessary to define an audience and focus the paper to
+this audience’s expertise.
+
+Specific Comments:
+
+As it was mentioned before, the main concern of the reviewer is the lack of focus in the paper.
+As such the main point of the review is to clarify who is the audience the paper is meant for and delve deeper into what is relevant for them.
+
+\ereview
+
+We feel that while the reviewer is right in that this manuscript
+targets more than one audience, and that we due to brevity reasons
+have to skip some details.
+However, we also feel that this is the point of a review, to give an overview for whoever wants to get an introduction to a subject.
+The details can be found in the publications we refer to.
+Regardless, we updated the text to better fit into a category of reader that we would define as a mass spectrometrist with a fair understanding of computational methods that is interested in knowing were the field of retention time predictions stands today.
+
+\breview
+
+Other than that, there are some points worth mentioning. They will be here presented in the order they appear on the manuscript.
+
+On section 2, it is written ``so-called linear gradients''. Why say ``so-called''? This expression gives the idea that the gradients in reality are not linear, but this isn’t further developed.
+
+\ereview
+We replaced the word so-called with setting the words linear gradient in italics, to indicate that a explanation of the term will follow. 
+\breview
+
+The reproducibility of RPLC is a big issue.
+Not only does a column get degraded with use, but there is also documented variability in products from different manufacturers.
+In this section it is also noted how the retention time depends on several experimental settings – ``variety of parameters such as column dimensions, the composition of stationary and mobile phase, temperature or gradient slope''.
+So we have several factors independent of the peptide sequence that influence greatly the retention time of a given peptide.
+In the following section it is said that ``given a chromatography setup, the
+sequence of a peptide should provide virtually all the information needed to predict its retention time''; however, given that there is so much variability in the chromatography setup in different labs and different experiments, we feel that it should be emphasized how big of an issue this is in retention time prediction and note efforts made in addressing this.
+
+\ereview
+We added a qualifying section to this section
+
+``However, in practice such predictive models is not as straight
+forward as it might seem. For instance, one important part of such a
+model would be the peptides three dimensional conformation. Although
+the current structure prediction methods are quite accurate in
+predicting three dimensional conformation of short amino acid chains,
+such methods will not be equipped with the flexibility to model the
+various conditions of a chromatographic experiment. For instance, many
+peptides are likely changing their confirmation as a function of the
+amount of organic solvent in the mobile phase, an effect that is
+likely to have an important contribution to each peptides retention
+time. Hence, most efforts to predict retention time have moved into
+treating the chromatographic process as a black box, that is a system
+that is only viewed in terms of its inputs an outputs.''
+
+\breview
+
+There are some issues with the caption for Fig. 2. It is also not clear if there are points that fall on the regression line. 
+\ereview
+
+We changed the formating of the manuscript so that the caption of Figure2 should stand clear. There is no regression line in Fig. 2.
+
+\breview
+
+Besides, this figure is not representative of a complex sample or of the reproducibility (or variability) in experiments between different labs, or even different machines.
+As such the figure can be misleading in terms of evaluating the reproducibility of a shotgun proteomics experiment.
+
+\ereview
+
+We added a qualifying sentence to the caption of the Figure: ``It
+should be noted that this relatively high reproducibility normally
+only appears between runs under near to identical experiments, and do
+not carry over into other experimental conditions or instrumentation.''
+
+\breview
+
+Across section 3, there is a lack of evaluation of the prediction methods.
+In fact at a certain point it is said that one is ``good'', one ``excellent'', and another that has ``lower'' accuracy followed by one that is ``reasonable''.
+It is inferred that if these qualitative comparisons are made by the authors, then there must be some accuracy measure available.
+It would be preferable to have a table with these values.
+\ereview
+
+We removed all these rather sloppy unqualified references to methods accuracy.
+
+\breview
+
+Table 1 falls out of the page.
+\ereview
+We modified the formating of the table.
+\breview
+
+In section 3.1, the last paragraph doesn’t seem relevant to the section it is included in. Maybe it would make more sense to include it in the applications section.
+\ereview
+
+We did not feel that the paragraph sat better in the application section and letted the sentence remain as is.
+
+\breview
+
+In section 3.2, it feels relevant to introduce the approach ELUDE takes regarding retention indexes.
+Even if they obtained via, are used as features for, a machine learning approach, it is still relevant to this section.
+\ereview
+
+We added a couple of sentences describing this.
+\breview
+
+Section 3.3 is rather brief. The reviewer feels there should be more information on these methods (or none at all).
+\ereview
+\breview
+
+On sections 3.1, 3.2 and 3.3 there is no evaluation of any of the introduced methods.
+In section 3.4 we see the sort of qualitative comparisons mentioned above.
+We note that there is no mention of how the performance of ELUDE compares to the other methods, even if qualitatively.
+\ereview
+\breview
+
+If the proposed audience to this paper are scientists applying machine learning methods to retention time prediction, more information on the features used and the methods themselves would be relevant.
+\ereview
+\breview
+
+In the discussion the reviewer feels that it should be discussed not only how retention time prediction can be improved, but also the limitations of how accurate predictions can be, given the issues mentioned above.
+Another relevant question is how much more relevant do these predictions need be given the applications enumerated in section 5.
+
 \ereview

### 2015-10-06 retention.tex
have to leave

@@
 is an essential aspect of any RPLC experiment. Currently, the most
-commonly used approach is based on so-called linear gradients, where
+commonly used approach is based on {\em linear gradients}, where
 the increase in solvent B follows a linear function. Most of the
@@
 virtually all the information needed to predict its retention time.
-However, in practice this is not as straight forward as it might seem, as it for instance is difficult to predict a peptide's three dimensional structure from its amino acid sequence.
 
-%\begin{wrapfigure}{r}{.45\textwidth}
-%\vspace{-10pt}
 \begin{figure}[htbp]
@@
   out. The retention times of the peptides confidently identified in
-  both runs are displayed.}
+  both runs are displayed.
+It should be noted that this relatively high reproducibility normally only appears between runs under near to identical experiments, and do not carry over into other experimental conditions or instrumentation.}
 \end{figure}
-%\vspace{-25pt}
-%\end{wrapfigure}
 
-% iRT? \cite{Escher:2012qv}
-Following these observations, a variety of retention time prediction
-methods has been proposed. Most naturally one could divide these
+However, in practice such predictive models is not as straight forward
+as it might seem. For instance, one important part of such a model
+would be the peptides three dimensional conformation. Although the
+current structure prediction methods are quite accurate in predicting
+three dimensional conformation of short amino acid chains, such
+methods will not be equipped with the flexibility to model the various
+conditions of a chromatographic experiment. For instance, many
+peptides are likely changing their confirmation as a function of the
+amount of organic solvent in the mobile phase, an effect that is
+likely to have an important contribution to each peptides retention
+time. Hence, most efforts to predict retention time have moved into
+treating the chromatographic process as a black box, that is a system
+that is only viewed in terms of its inputs an outputs.
+
+Following these observations,a variety of retention time prediction methods has been proposed. Most naturally one could divide these
 methods into four classes: look-up tables-based approaches,
@@
  \label{tab:rtmethods}
- \begin{tabular}{lll}
+ \begin{tabular}{lp{9cm}l}
 
@@
 % ANN & Link? & \cite{petritis2003,petritis2006improved} \\
-RTModel, RTPredict &  \url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTPredict.html}, 
+RTPredict &  \url{ http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTPredict.html}, 
  & \cite{rtpredict, rtpredictImproved} \\
-&\url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTModel.html}  & \\
+RTModel &\url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTModel.html}  & \cite{rtpredict, rtpredictImproved} \\
 
 {\sc Elude} & \url{http://elude.sbc.su.se/} & \cite{elude1, elude2} \\
-iRT-calculator & http://www.biognosys.ch/products/irt-calculator.html
-& \cite{irt} \\
+iRT-calculator & \url{http://www.biognosys.ch/products/irt-calculator.html} & \cite{irt} \\
 \hline
@@
 diverging from the setup on which the algorithm was optimized
-\cite{Spicer2007}.
+\cite{Spicer2007}. Consequently, more advanced methods such as {\sc
+Elude} \cite{elude1,elude2} estimates retention time indices for each
+dataset as hand. In these cases the indeces are used to generate
+features that are used as an input for final retention time
+estimates. We will come back to this in section \ref{sec:ML}.
 
@@
 peptides with known retention times and evaluated using a set of
-$\sim$5000 peptides originating from another organism. The algorithm
-reported good accuracy, especially taking into account that it only
-used information about the amino acid composition of the peptides.
+$\sim$5000 peptides originating from another organism.
 
@@
 peptides identified in more than 12,000 shotgun experiments. The
-trained algorithm reported excellent performance for the 1303 peptides
-on which it was evaluated.
+trained algorithm reported better performance for the 1303 peptides
+on which it was evaluated, than its predecesor.
 
@@
 setups is commonly used across research laboratories. To address this,
-Klammer {\em et al.} proposed a predictor based on support vector machines
-(SVMs) \cite{klammer2007improving}. Their method required fewer
-training peptides and could thus be easily adapted to new
-chromatographic conditions, but yielded lower prediction
-accuracy. Subsequently, a more elaborate SVM-based method using a
-customized kernel function was proposed \cite{rtpredict, rtpredictImproved}. This
-approach provided reasonable prediction accuracy, while requiring as
-little as 50 peptides for training.
+Klammer {\em et al.} proposed a predictor based on support vector
+machines (SVMs) \cite{klammer2007improving}. Their method required
+fewer training peptides and could thus be easily adapted to new
+chromatographic conditions, but yielded lower prediction accuracy.
+Subsequently, a more elaborate SVM-based method using a customized
+kernel function was proposed \cite{rtpredict, rtpredictImproved}. This
+approach provided an algorithm that required as little as 50 peptides
+for training.
 
@@
 probably additional factors are involved, and further studies are
-required to reach the desired level of accuracy.
+required to even understand the error sources in the process.
 
@@
 
-Palmblad {\em et al.} showed that matching theoretical and observed masses
-and retention times provides a simple, yet powerful method to identify
-peptides in simple biological
-mixtures \cite{palmblad2002prediction}. Strittmatter {\em et al.} used
+Palmblad {\em et al.} showed that matching theoretical and observed
+masses and retention times provides a simple, yet powerful method to
+identify peptides in simple biological mixtures
+\cite{palmblad2002prediction}. Strittmatter {\em et al.} used
 retention time as a parameter in a discriminant function that
-differentiated between correct and incorrect peptide
-identifications \cite{Strittmatter2004}. Following this approach, they
-reported $\sim$9\% more confident peptide identifications in a typical
-shotgun experiment. Along the same lines, Klammer {\em et al.} demonstrated
+differentiated between correct and incorrect peptide identifications
+\cite{Strittmatter2004}. Following this approach, they reported
+$\sim$9\% more confident peptide identifications in a typical shotgun
+experiment. Along the same lines, Klammer {\em et al.} demonstrated
 that for specific experimental conditions, just filtering out peptides
 with large deviations between observed and predicted retention time
-leads to up to 50\% more
-identifications \cite{klammer2007improving}.
+leads to up to 50\% more identifications \cite{klammer2007improving}.
 

### 2015-10-07 retention.tex
Auto

@@
 inference procedure is to also predict properties such as their
-iso-electric point~\cite{branca2014hirief} or of particular interest
+iso-electric point~\cite{perez2011silico,branca2014hirief} or of particular interest
 for this review, chromatographic retention

### 2015-10-07 retention.tex
some new answers

--- 2015-10-06 retention.tex
@@
 inference procedure is to also predict properties such as their
-iso-electric point~\cite{branca2014hirief} or of particular interest
-for this review, chromatographic retention
-time \cite{cerqueira2010mude}.
+iso-electric point~\cite{perez2011silico,branca2014hirief} or of
+particular interest for this review, chromatographic retention time
+\cite{cerqueira2010mude}. One way to determie a predicatble sparation
+techniques usefulness is to estimate its ability to remove candidate
+peptides of any detected feature, normally measured by a false
+positive ratio (FPR), also known as the methods fall-out. Current
+retention time prediction algorithms have an FPR of about
+$\frac{1}{5}$, meaning that it can remove about four out of five
+candidate peptides for a given feature. This could be compared to the
+FPR of about $\frac{1}{200}$ for predictions of high-resolution
+iso-electric focusing experiments~\cite{branca2014hirief}. The
+retention time prediction have toimprove quite a bit to come to level
+with the fall-out of iso-electric focusing, however, almost all mass
+spectrometry-based proteomics use reversed phase chromatography, while
+very few studies are using iso-electric focusing.
+
+% An estimate of a FPR for MS1's would be great here.
 

### 2015-10-08 response.tex
some new answers
answered all the questions

@@
 
-The authors give a good overview of the existing methods to predict the retention time.
-However, the description of the methods is sometimes too brief, and slightly biased towards their own developed methods.
-Extending the descriptions of the methods that are not developed by the authors should be considered.
-The applications and discussion section are an added value to the paper.
+The authors give a good overview of the existing methods to predict
+the retention time.  However, the description of the methods is
+sometimes too brief, and slightly biased towards their own developed
+methods. Extending the descriptions of the methods that are not
+developed by the authors should be considered. The applications and
+discussion section are an added value to the paper.
 
@@
 
-We would like to thank the reviewer for the positive feedback. We have revised the sections were the other methods contribution is discussed
+We would like to thank the reviewer for the positive feedback. We have
+revised the sections were the other methods contribution is discussed
 
@@
 
-As it was mentioned before, the main concern of the reviewer is the lack of focus in the paper.
-As such the main point of the review is to clarify who is the audience the paper is meant for and delve deeper into what is relevant for them.
+As it was mentioned before, the main concern of the reviewer is the
+lack of focus in the paper. As such the main point of the review is to
+clarify who is the audience the paper is meant for and delve deeper
+into what is relevant for them.
 
@@
 targets more than one audience, and that we due to brevity reasons
-have to skip some details.
-However, we also feel that this is the point of a review, to give an overview for whoever wants to get an introduction to a subject.
-The details can be found in the publications we refer to.
-Regardless, we updated the text to better fit into a category of reader that we would define as a mass spectrometrist with a fair understanding of computational methods that is interested in knowing were the field of retention time predictions stands today.
+have to skip some details. However, we also feel that this is the
+point of a review, to give an overview for whoever wants to get an
+introduction to a subject. The details can be found in the
+publications we refer to. Regardless, we updated the text to better
+fit into a category of reader that we would define as a mass
+spectrometrist with a fair understanding of computational methods that
+is interested in knowing were the field of retention time predictions
+stands today.
 
@@
 
-Other than that, there are some points worth mentioning. They will be here presented in the order they appear on the manuscript.
+Other than that, there are some points worth mentioning. They will be
+here presented in the order they appear on the manuscript.
 
-On section 2, it is written ``so-called linear gradients''. Why say ``so-called''? This expression gives the idea that the gradients in reality are not linear, but this isn’t further developed.
+On section 2, it is written ``so-called linear gradients''. Why say
+``so-called''? This expression gives the idea that the gradients in
+reality are not linear, but this isn’t further developed.
 
 \ereview
-We replaced the word so-called with setting the words linear gradient in italics, to indicate that a explanation of the term will follow. 
+
+This is in part due to the strange nomenclature. In other fields a
+gradient refers to the vector of spacial partial derivates, not a time
+derivate.
+
+However, to avoid confusion, we replaced the word so-called with
+setting the words linear gradient in italics, to indicate that a
+explanation of the term will follow.
+
 \breview
 
-The reproducibility of RPLC is a big issue.
-Not only does a column get degraded with use, but there is also documented variability in products from different manufacturers.
-In this section it is also noted how the retention time depends on several experimental settings – ``variety of parameters such as column dimensions, the composition of stationary and mobile phase, temperature or gradient slope''.
-So we have several factors independent of the peptide sequence that influence greatly the retention time of a given peptide.
-In the following section it is said that ``given a chromatography setup, the
-sequence of a peptide should provide virtually all the information needed to predict its retention time''; however, given that there is so much variability in the chromatography setup in different labs and different experiments, we feel that it should be emphasized how big of an issue this is in retention time prediction and note efforts made in addressing this.
+The reproducibility of RPLC is a big issue. Not only does a column get
+degraded with use, but there is also documented variability in
+products from different manufacturers. In this section it is also
+noted how the retention time depends on several experimental settings
+– ``variety of parameters such as column dimensions, the composition
+of stationary and mobile phase, temperature or gradient slope''. So we
+have several factors independent of the peptide sequence that
+influence greatly the retention time of a given peptide. In the
+following section it is said that ``given a chromatography setup, the
+sequence of a peptide should provide virtually all the information
+needed to predict its retention time''; however, given that there is
+so much variability in the chromatography setup in different labs and
+different experiments, we feel that it should be emphasized how big of
+an issue this is in retention time prediction and note efforts made in
+addressing this.
 
 \ereview
-We added a qualifying section to this section
+We added a qualifying section to this section.
 
@@
 
-There are some issues with the caption for Fig. 2. It is also not clear if there are points that fall on the regression line. 
+There are some issues with the caption for Fig. 2. It is also not
+clear if there are points that fall on the regression line.
+
 \ereview
 
-We changed the formating of the manuscript so that the caption of Figure2 should stand clear. There is no regression line in Fig. 2.
+We changed the formating of the manuscript so that the caption of
+Figure2 should stand clear. There is no regression line in Fig. 2.
 
@@
 
-Besides, this figure is not representative of a complex sample or of the reproducibility (or variability) in experiments between different labs, or even different machines.
-As such the figure can be misleading in terms of evaluating the reproducibility of a shotgun proteomics experiment.
+Besides, this figure is not representative of a complex sample or of
+the reproducibility (or variability) in experiments between different
+labs, or even different machines. As such the figure can be misleading
+in terms of evaluating the reproducibility of a shotgun proteomics
+experiment.
 
@@
 
-Across section 3, there is a lack of evaluation of the prediction methods.
-In fact at a certain point it is said that one is ``good'', one ``excellent'', and another that has ``lower'' accuracy followed by one that is ``reasonable''.
-It is inferred that if these qualitative comparisons are made by the authors, then there must be some accuracy measure available.
-It would be preferable to have a table with these values.
+Across section 3, there is a lack of evaluation of the prediction
+methods. In fact at a certain point it is said that one is ``good'',
+one ``excellent'', and another that has ``lower'' accuracy followed by
+one that is ``reasonable''. It is inferred that if these qualitative
+comparisons are made by the authors, then there must be some accuracy
+measure available. It would be preferable to have a table with these
+values.
+
 \ereview
 
-We removed all these rather sloppy unqualified references to methods accuracy.
+We removed all these unqualified references to methods accuracy.
 
@@
 \ereview
+
+We added some more details on how BioLCCC works to the text.
+
 \breview
 
-On sections 3.1, 3.2 and 3.3 there is no evaluation of any of the introduced methods.
-In section 3.4 we see the sort of qualitative comparisons mentioned above.
-We note that there is no mention of how the performance of ELUDE compares to the other methods, even if qualitatively.
+On sections 3.1, 3.2 and 3.3 there is no evaluation of any of the
+introduced methods. In section 3.4 we see the sort of qualitative
+comparisons mentioned above. We note that there is no mention of how
+the performance of ELUDE compares to the other methods, even if
+qualitatively.
+
 \ereview
+We added a reference to a benchmark to the section.
 \breview
 
-If the proposed audience to this paper are scientists applying machine learning methods to retention time prediction, more information on the features used and the methods themselves would be relevant.
-\ereview
-\breview
-
-In the discussion the reviewer feels that it should be discussed not only how retention time prediction can be improved, but also the limitations of how accurate predictions can be, given the issues mentioned above.
-Another relevant question is how much more relevant do these predictions need be given the applications enumerated in section 5.
+If the proposed audience to this paper are scientists applying machine
+learning methods to retention time prediction, more information on the
+features used and the methods themselves would be relevant.
 
@@
 
+We included an brief enumeration of some of the features used in Elude.
+
+\breview
+
+In the discussion the reviewer feels that it should be discussed not
+only how retention time prediction can be improved, but also the
+limitations of how accurate predictions can be, given the issues
+mentioned above. Another relevant question is how much more relevant
+do these predictions need be given the applications enumerated in
+section 5.
+
+\ereview
+
+Estimate of how good predictions should be to be useful are generally
+hard to make. Better performance will always be more useful than worse
+performance.  Instead we added a comparison of the FPR of retention
+time predictions versus predictions of iso-electric point calculations.
+
 \end{document}

### 2015-10-08 retention.tex
answered all the questions

@@
 % for a few other column
-Following these observations, Krokhin {\em et al.} developed SSRCalc
-\cite{Krokhin2004}, which is currently the most widely used
+Following these observations, Krokhin {\em et al.} developed {\sc
+SSRCalc} \cite{Krokhin2004}, which is currently the most widely used
 index-based retention time predictor. In the initial version of the
@@
 capture the chemical behavior of peptides on the chromatography
-column. One such method is BioLCCC, which attempts to model the
-connectivity of the amino acids in a peptide chain and the general
-mechanisms of chromatographic
-separation \cite{gorshkov2006}. Nevertheless, given the numerous
-factors involved in peptide separation, modeling approaches often
-involve a considerable number of simplifications.
+column. One such method is BioLCCC, models the chromatography process
+using a lattice based random walk model for the peptides, which are
+modeled as chains of monomers. The model also includes an entropy and
+energy compensation for the amount of peptides residing on the column 
+and a model of the interaction energies between the peptide and the
+stationary phase.
+
+Nevertheless, given the numerous factors involved in peptide
+separation, modeling approaches often involve a considerable number of
+simplifications.
 
@@
 chromatographic conditions, but yielded lower prediction accuracy.
-Subsequently, a more elaborate SVM-based method using a customized
-kernel function was proposed \cite{rtpredict, rtpredictImproved}. This
-approach provided an algorithm that required as little as 50 peptides
-for training.
+Subsequently, a more elaborate SVM-based method, {\sc RTPredict}, that
+ use a customized kernel function was proposed \cite{rtpredict,
+rtpredictImproved}. This approach provided an algorithm that required
+as little as 50 peptides for training.
 
@@
 
-Moruz {\em et al.} \cite{elude1} developed a software package for
+Moruz {\em et al.}~\cite{elude1} developed a software package for
 retention time prediction named {\sc Elude}. One of the main novelty
@@
 both a general hydrophobicity index, as well as a retention index
-trained for each dataset at hand. This ensured that the algorithm
-could be applied to different chromatography setups with minimal loss
-in performance. For cases when little training data was available,
-{\sc Elude} implemented an additional workflow based on a library of
+trained for each dataset at hand. The features included properties
+such as average hydrophobicity, the N- and C-termini's hydrophobicity,
+the stretch of the five most and least hydrophobic amino acids,
+helicity, and the number of occurrences of the 25\% most or least
+hydrophobic amino acids. This ensured that the algorithm could be
+applied to different chromatography setups with minimal loss in
+performance. For cases when little training data was available, {\sc
+Elude} implemented an additional workflow based on a library of
 pre-trained retention models. This workflow required only a few tens
@@
 
-However, despite all these efforts, it should be pointed out that none
-of the retention time prediction methods currently available is able
-to provide a prediction precision comparable to the small shifts
-observed when comparing retention time of peptides from two technical
-replicates (Figure \ref{fig:repr}). Recently, the lack of suitable
+In a benchmark based on the merit of how large retention time window,
+measured as a fraction of the gradient time was be needed to include
+95\% of the peptides of a series of 14 runs, {\sc Elude} outperformed
+{\sc SSRCalc} and {\sc RTPredict}, needing only on average 22\% as
+compared to 28\% and 26\% respectively~\cite{elude1}. However, despite
+all these efforts, it should be pointed out that none of the retention
+time prediction methods currently available is able to provide a
+prediction precision comparable to the small shifts observed when
+comparing retention time of peptides from two technical replicates
+(Figure \ref{fig:repr}). Recently, the lack of suitable
 representations for secondary structure elements was indicated as one
@@
 Moreover, targeted proteomics experiments, such as Selected Reaction
-Monitoring (SRM), are dependent on the predictability of a
-peptide's behavior. In such experiments the retention times and the
-suitable transitions are needed as input to the analysis. Frequently,
-such information is derived from previous shotgun analysis. As the
-capacity of targeted proteomics experiments expands by the
-introduction of the more comprehensive data independent acquisition
-techniques \cite{Venable2004} (a.k.a. SWATH analysis), such
-pre-identification of peptide transitions and retention times
-threatens to become a limiting factor. Our current methods to identify
-peptides from Data Independent Acquisition are only as comprehensive as the previous shotgun
-analysis would be. An attractive alternative is to predict retention
-time and transition using {\em de novo} predictions from the peptides
-amino acid sequences.  Such a system of identifications not dependent
-on previous peptide identifications from shotgun proteomics experiments
+Monitoring (SRM), are dependent on the predictability of a peptide's
+behavior. In such experiments the retention times and the suitable
+transitions are needed as input to the analysis. Frequently, such
+information is derived from previous shotgun analysis. As the capacity
+of targeted proteomics experiments expands by the introduction of the
+more comprehensive data independent acquisition techniques
+\cite{Venable2004} (a.k.a. SWATH analysis), such pre-identification of
+peptide transitions and retention times threatens to become a limiting
+factor. Our current methods to identify peptides from Data Independent
+Acquisition are only as comprehensive as the previous shotgun analysis
+would be. An attractive alternative is to predict retention time and
+transition using {\em de novo} predictions from the peptides amino
+acid sequences.  Such a system of identifications not dependent on
+previous peptide identifications from shotgun proteomics experiments
 would lead to a more direct experimental design and hopefully to more

### 2015-10-12 retention.tex
Submitted version

@@
 \usepackage{multirow}
+%\usepackage[nomarkers,figuresonly]{endfloat}
+\usepackage[nomarkers,nolists,noheads]{endfloat}
 
@@
 applications of retention time predictions.
+
 % Section~\ref{sec:irt} briefly introduces the use of peptide retention
@@
 
-\begin{figure}[!h]
+\begin{figure}[p]
 \centering
@@
 
-\begin{figure}[htbp]
+\begin{figure}[p]
 \centering
@@
 
-\begin{figure}[!h]
+\begin{figure}[p]
 \centering \includegraphics[clip=true,
@@
 
-\begin{table}
+\begin{table}[p]
  \caption{{\bf Retention time prediction methods.} We
@@
 Elude} \cite{elude1,elude2} estimates retention time indices for each
-dataset as hand. In these cases the indeces are used to generate
+dataset as hand. In these cases the indices are used to generate
 features that are used as an input for final retention time
@@
 trained algorithm reported better performance for the 1303 peptides
-on which it was evaluated, than its predecesor.
+on which it was evaluated, than its predecessor.
 
@@
 
-\begin{figure}[!h]
+\begin{figure}[p]
 \centering 
@@
 inference procedure is to also predict properties such as their
-iso-electric point~\cite{perez2011silico,branca2014hirief} or of
+isoelectric point~\cite{perez2011silico,branca2014hirief} or of
 particular interest for this review, chromatographic retention time
-\cite{cerqueira2010mude}. One way to determie a predicatble sparation
-techniques usefulness is to estimate its ability to remove candidate
-peptides of any detected feature, normally measured by a false
-positive ratio (FPR), also known as the methods fall-out. Current
-retention time prediction algorithms have an FPR of about
-$\frac{1}{5}$, meaning that it can remove about four out of five
-candidate peptides for a given feature. This could be compared to the
-FPR of about $\frac{1}{200}$ for predictions of high-resolution
+\cite{cerqueira2010mude}.
+
+One way to determine a predictable separation techniques usefulness is
+to estimate its ability to remove candidate peptides of any detected
+feature, normally measured by a false positive ratio (FPR), also known
+as the methods fall-out. The FPR is of course a function of the
+selected thresholds, but if we select a treshold were say $\sim$95\%
+of the known examples get correctly included, what FPR do we find at
+such level?  Current retention time prediction algorithms have an FPR
+of about $\frac{1}{5}$, meaning that it can remove about four out of
+five candidate peptides for a given feature. This could be compared to
+the FPR of about $\frac{1}{200}$ for predictions of high-resolution
 iso-electric focusing experiments~\cite{branca2014hirief}. The
-retention time prediction have toimprove quite a bit to come to level
-with the fall-out of iso-electric focusing, however, almost all mass
+retention time prediction have to improve quite a bit to come to level
+with the fall-out of isoelectric focusing, however, almost all mass
 spectrometry-based proteomics use reversed phase chromatography, while
@@
 %%  LocalWords:  FRActional DIagonal COFRADIC datasets Palmblad
-%%  LocalWords:  Strittmatter unfragmented
+%%  LocalWords:  Strittmatter unfragmented acetonitrile chromatogram

### 2015-11-09 retention.tex
dirks edits

@@
 development of efficient algorithms for data interpretation a major
-area within proteomics research. A particular difficult challenge in
+area within proteomics research. A particularly difficult challenge in
 this context is posed by the tremendous complexity of the data.
@@
 posttranslational modifications~\cite{Lemeer2009}. To manage this
-diversity, increasing efforts are made to extract besides the
-fragmentation spectra additional information from each MS experiment.
+diversity, increasing efforts are made to extract additional
+information besides the fragmentation spectra from each MS experiment.
 
@@
 
-However, in practice such predictive models is not as straight forward
-as it might seem. For instance, one important part of such a model
-would be the peptides three dimensional conformation. Although the
-current structure prediction methods are quite accurate in predicting
-three dimensional conformation of short amino acid chains, such
-methods will not be equipped with the flexibility to model the various
-conditions of a chromatographic experiment. For instance, many
-peptides are likely changing their confirmation as a function of the
-amount of organic solvent in the mobile phase, an effect that is
-likely to have an important contribution to each peptides retention
-time. Hence, most efforts to predict retention time have moved into
-treating the chromatographic process as a black box, that is a system
-that is only viewed in terms of its inputs an outputs.
+However, in practice such predictive models are not as straight
+forward to design as it might seem. For instance, one important part
+of such a model would be the peptides three dimensional
+conformation. Although the current structure prediction methods are
+quite accurate in predicting three dimensional conformation of short
+amino acid chains, such methods will not be equipped with the
+flexibility to model the various conditions of a chromatographic
+experiment. For instance, many peptides are likely changing their
+confirmation as a function of the amount of organic solvent in the
+mobile phase, an effect that is likely to have an important
+contribution to each peptides retention time. Hence, most efforts to
+predict retention time have moved into treating the chromatographic
+process as a black box, that is, a system that is only viewed in terms
+of its inputs and outputs.
 
-Following these observations,a variety of retention time prediction methods has been proposed. Most naturally one could divide these
+Following these observations, a variety of retention time prediction
+methods have been proposed. Most naturally one could divide these
 methods into four classes: look-up tables-based approaches,
@@
 \cite{Spicer2007}. Consequently, more advanced methods such as {\sc
-Elude} \cite{elude1,elude2} estimates retention time indices for each
+  Elude} \cite{elude1,elude2} estimate retention time indices for each
 dataset as hand. In these cases the indices are used to generate
@@
 capture the chemical behavior of peptides on the chromatography
-column. One such method is BioLCCC, models the chromatography process
-using a lattice based random walk model for the peptides, which are
-modeled as chains of monomers. The model also includes an entropy and
-energy compensation for the amount of peptides residing on the column 
-and a model of the interaction energies between the peptide and the
-stationary phase.
+column. One such method is BioLCCC, which models the chromatography
+process using a lattice based random walk model for the peptides. The
+method models peptides as chains of monomers. The model also includes
+an entropy and energy compensation for the amount of peptides residing
+on the column and a model of the interaction energies between the
+peptide and the stationary phase.
 
@@
 Subsequently, a more elaborate SVM-based method, {\sc RTPredict}, that
- use a customized kernel function was proposed \cite{rtpredict,
-rtpredictImproved}. This approach provided an algorithm that required
-as little as 50 peptides for training.
+uses a customized kernel function was proposed \cite{rtpredict,
+  rtpredictImproved}. This approach provided an algorithm that
+required as little as 50 peptides for training.
 
@@
 
-In a benchmark based on the merit of how large retention time window,
-measured as a fraction of the gradient time was be needed to include
-95\% of the peptides of a series of 14 runs, {\sc Elude} outperformed
-{\sc SSRCalc} and {\sc RTPredict}, needing only on average 22\% as
-compared to 28\% and 26\% respectively~\cite{elude1}. However, despite
-all these efforts, it should be pointed out that none of the retention
-time prediction methods currently available is able to provide a
-prediction precision comparable to the small shifts observed when
-comparing retention time of peptides from two technical replicates
-(Figure \ref{fig:repr}). Recently, the lack of suitable
-representations for secondary structure elements was indicated as one
-of the main causes of prediction errors \cite{Reimer2012}.  Most
-probably additional factors are involved, and further studies are
-required to even understand the error sources in the process.
+In a benchmark based on the merit of how large the retention time
+window, measured as a fraction of the gradient time was needed to
+include 95\% of the peptides of a series of 14 runs, {\sc Elude}
+outperformed {\sc SSRCalc} and {\sc RTPredict}, needing only on
+average 22\% as compared to 28\% and 26\%
+respectively~\cite{elude1}. However, despite all these efforts, it
+should be pointed out that none of the retention time prediction
+methods currently available is able to provide a prediction precision
+comparable to the small shifts observed when comparing retention time
+of peptides from two technical replicates (Figure
+\ref{fig:repr}). Recently, the lack of suitable representations for
+secondary structure elements was indicated as one of the main causes
+of prediction errors \cite{Reimer2012}.  Most probably additional
+factors are involved, and further studies are required to even
+understand the error sources in the process.
 
@@
 peptides as well. This was straightforward to achieve once the authors
-observed that many of their predictive features of their method were
-all functions of the peptides amino acid sequence and a retention
+observed that many of the predictive features of their method were all
+functions of the peptide's amino acid sequence and a retention
 index. As such retention indices are estimated for each dataset at
@@
 modified amino acid (Figure \ref{fig:elude}). The new version of {\sc
-ELUDE} was evaluated on peptides containing artifactual modifications
-introduced as part of an MS protocol called COmbined FRActional
-DIagonal Chromatography (COFRADIC) \cite{Gevaert2002}. The results
-showed that {\sc ELUDE} yielded equally accurate predictions for
-modified and unmodified peptides \cite{elude2}.
+  ELUDE} was evaluated on peptides containing artifactual
+modifications introduced as part of an MS protocol called COmbined
+FRActional DIagonal Chromatography (COFRADIC) \cite{Gevaert2002}. The
+results showed that {\sc ELUDE} yielded equally accurate predictions
+for modified and unmodified peptides \cite{elude2}.
 
@@
 to estimate its ability to remove candidate peptides of any detected
-feature, normally measured by a false positive ratio (FPR), also known
-as the methods fall-out. The FPR is of course a function of the
-selected thresholds, but if we select a treshold were say $\sim$95\%
-of the known examples get correctly included, what FPR do we find at
-such level?  Current retention time prediction algorithms have an FPR
-of about $\frac{1}{5}$, meaning that it can remove about four out of
-five candidate peptides for a given feature. This could be compared to
-the FPR of about $\frac{1}{200}$ for predictions of high-resolution
+feature. This can be measured by a false positive ratio (FPR), also
+known as the method's fall-out. The FPR is of course a function of the
+selected thresholds, but if we select a threshold where $\sim$95\% of
+the known examples get correctly included, what FPR do we find at such
+level?  Current retention time prediction algorithms have an FPR of
+about $\frac{1}{5}$, meaning that it can remove about four out of five
+candidate peptides for a given feature. This could be compared to the
+FPR of about $\frac{1}{200}$ for predictions of high-resolution
 iso-electric focusing experiments~\cite{branca2014hirief}. The
-retention time prediction have to improve quite a bit to come to level
+retention time prediction has to improve quite a bit to come to level
 with the fall-out of isoelectric focusing, however, almost all mass

### 2015-11-09 retention.tex
edits from word grammar

@@
 % say something about gradient and washing 
-Since the elution of the peptides is determined by the proportion of
-solvent B present at each time point, the scheme used to add solvent B
-is an essential aspect of any RPLC experiment. Currently, the most
-commonly used approach is based on {\em linear gradients}, where
-the increase in solvent B follows a linear function. Most of the
-peptides elute before the percentage of solvent B reaches 40\%, and
-thus gradients often go from 2 to 40\% B. A high increase in
-percentage B at the end of the run ensures that all the remaining
-compounds are washed off from the column.
+Since the proportion of solvent B determines the elution of peptides,
+the scheme used to add solvent B is an essential aspect of any RPLC
+experiment. Currently, the most commonly used approach is based on
+{\em linear gradients}, where the increase in solvent B follows a
+linear function. Most of the peptides elute before the percentage of
+solvent B reaches 40\%, and thus gradients often go from 2 to 40\%
+B. A high increase in percentage B at the end of the run ensures that
+all the remaining compounds are washed off from the column.
 
@@
 However, in practice such predictive models are not as straight
-forward to design as it might seem. For instance, one important part
-of such a model would be the peptides three dimensional
+forward to design, as it might seem. For instance, one important part
+of such a model would be the peptides three-dimensional
 conformation. Although the current structure prediction methods are
-quite accurate in predicting three dimensional conformation of short
+quite accurate in predicting three-dimensional conformation of short
 amino acid chains, such methods will not be equipped with the
@@
 
-A limitation of index-based methods are that they are optimized for
+A limitation of index-based methods is that they are optimized for
 predicting retention times for certain chromatography setups. Thus,
@@
 
-In a follow-up work, the method was extended by representing each
-peptide by 1052 instead of 20 features
-\cite{petritis2006improved}. The new descriptors encoded the identity
-of each amino acid at each position in a sequence, as well as the
-length and the hydrophobic moment of the peptide. The new neural
-network was trained on more than 300,000 peptide sequences. To obtain
-such a large number of peptides with known retention times, a strategy
-to normalize elution times was developed. This allowed pooling of
-peptides identified in more than 12,000 shotgun experiments. The
-trained algorithm reported better performance for the 1303 peptides
-on which it was evaluated, than its predecessor.
+The method was extended by representing each peptide by 1052 instead
+of 20 features in a follow-up study~\cite{petritis2006improved}. The
+new descriptors encoded the identity of each amino acid at each
+position in a sequence, as well as the length and the hydrophobic
+moment of the peptide. The new neural network was trained on more than
+300,000 peptide sequences. To obtain such a large number of peptides
+with known retention times, a strategy to normalize elution times was
+developed. This allowed pooling of peptides identified in more than
+12,000 shotgun experiments. The trained algorithm reported better
+performance for the 1303 peptides on which it was evaluated, than its
+predecessor.
 
@@
 little efforts have been directed toward investigating the
-chromatographic behavior of posttranslationally modified peptides. The
-few studies available so far have mostly focused on phosphorylated
-peptides \cite{Kim2007}, or a couple of artifactual modifications
-commonly introduced during sample preparation
+chromatographic behavior of post-translationally modified
+peptides. The few studies available so far have mostly focused on
+phosphorylated peptides \cite{Kim2007}, or a couple of artifactual
+modifications commonly introduced during sample preparation
 \cite{Reimer2011}. Consequently, few retention time prediction methods
@@
 one can structure the prediction into multiple layers. The drawback of
-such models are that they require large training sets. Given the very
+such models is that they require large training sets. Given the very
 large sets of peptides with retention times that are assembled in
@@
 
-One way to determine a predictable separation techniques usefulness is
+One way to determine a predictable separation technique's usefulness is
 to estimate its ability to remove candidate peptides of any detected
@@
 FPR of about $\frac{1}{200}$ for predictions of high-resolution
-iso-electric focusing experiments~\cite{branca2014hirief}. The
+isoelectric focusing experiments~\cite{branca2014hirief}. The
 retention time prediction has to improve quite a bit to come to level

### 2015-11-11 retention.tex
Antecedents to this etc.

@@
 Today a majority of the mass spectrometry (MS) protocols in use have a
-high-throughput, generating extensive amounts of data. This has
-brought up a whole new set of analytical challenges, making the
-development of efficient algorithms for data interpretation a major
-area within proteomics research. A particularly difficult challenge in
-this context is posed by the tremendous complexity of the data.
-Proteins display a wide range of physico-chemical properties and are
-often found at very different amounts in complex samples, spanning as
-much as 12 orders of magnitude in relative abundance~\cite{Angel2012}.
-Furthermore, a large proportion of proteins are subjected to
-posttranslational modifications~\cite{Lemeer2009}. To manage this
-diversity, increasing efforts are made to extract additional
-information besides the fragmentation spectra from each MS experiment.
+high-throughput, generating extensive amounts of data. This abundance
+of data has brought up a whole new set of analytical challenges,
+making the development of efficient algorithms for data interpretation
+a major area within proteomics research. A particularly difficult
+challenge in this context is posed by the tremendous complexity of the
+data.  Proteins display a wide range of physico-chemical properties
+and are often found at very different amounts in complex samples,
+spanning as much as 12 orders of magnitude in relative
+abundance~\cite{Angel2012}.  Furthermore, a large proportion of
+proteins are subjected to posttranslational
+modifications~\cite{Lemeer2009}. To manage this diversity, increasing
+efforts are made to extract additional information besides the
+fragmentation spectra from each MS experiment.
 
@@
 to reduce the number of peptides simultaneously injected into the mass
-spectrometer. This increases the probability of each individual
-peptide to be selected for fragmentation, which in its turn, increases
-the proteome coverage achieved in an experiment.  However, besides
-this, the separation process in itself provides useful information
-about the analytes. On the one hand, the instrument always records the
-time when a peptide elutes from the chromatography column, {\em i.e.}
-the retention time. On the other hand, the retention time can also be
-estimated by examining the sequence of the peptide. The availability
-of both predicted and observed retention times has proven to be a
-powerful feature, with several applications in MS-based proteomics
-\cite{Palmblad2013}.
+spectrometer. This reduction in complexity increases the probability
+of each individual peptide to be selected for fragmentation, which in
+its turn, increases the proteome coverage achieved in an experiment.
+However, the separation process also in itself provides useful
+information about the analytes. On the one hand, the instrument always
+records the time when a peptide elutes from the chromatography column,
+{\em i.e.}  the retention time. On the other hand, the retention time
+can also be estimated by examining the sequence of the peptide. The
+availability of both predicted and observed retention times has proven
+to be a powerful feature, with several applications in MS-based
+proteomics \cite{Palmblad2013}.
 
@@
 the preferred method for separating peptides prior to mass
-spectrometry analysis \cite{Xie201278}. This is mainly because the
-method provides good resolution, gives reproducible results, and most
-of all, is compatible with electrospray ionization. In virtually all
-modern mass spectrometers the RPLC system can be easily connected to
-the ionization source, such that the peptides can be automatically
-transferred to mass spectrometry analysis as they elute from the
-chromatography column.
+spectrometry analysis \cite{Xie201278}. This preference is mainly
+because the method provides good resolution, gives reproducible
+results, and most of all, is compatible with electrospray
+ionization. In virtually all modern mass spectrometers the RPLC system
+can be easily connected to the ionization source, such that the
+peptides can be automatically transferred to mass spectrometry
+analysis as they elute from the chromatography column.
 
@@
 % say something about gradient and washing 
-Since the proportion of solvent B determines the elution of peptides,
-the scheme used to add solvent B is an essential aspect of any RPLC
-experiment. Currently, the most commonly used approach is based on
-{\em linear gradients}, where the increase in solvent B follows a
-linear function. Most of the peptides elute before the percentage of
-solvent B reaches 40\%, and thus gradients often go from 2 to 40\%
-B. A high increase in percentage B at the end of the run ensures that
-all the remaining compounds are washed off from the column.
+Because the proportion of solvent B determines the elution of
+peptides, the scheme used to add solvent B is an essential aspect of
+any RPLC experiment. Currently, the most commonly used approach is
+based on {\em linear gradients}, where the increase in solvent B
+follows a linear function. Most of the peptides elute before the
+percentage of solvent B reaches 40\%, and thus gradients often go from
+2 to 40\% B. A high increase in percentage B at the end of the run
+ensures that all the remaining compounds are washed off from the
+column.
 
@@
 when a sample is reanalyzed under identical experimental conditions
-(Figure~\ref{fig:repr}). This suggests that peptides engage in
-well-defined interactions during separation, which depend only on the
-experimental setup employed and the properties of the peptides
-themselves. However, the properties of the peptides are determined to
-a large extent by their amino-acid sequence. As a consequence, given a
-chromatography setup, the sequence of a peptide should provide
-virtually all the information needed to predict its retention time.
+(Figure~\ref{fig:repr}). This consistency in elution time suggests
+that peptides engage in well-defined interactions during separation,
+which depend only on the experimental setup employed and the
+properties of the peptides themselves. However, the properties of the
+peptides are determined to a large extent by their amino-acid
+sequence. As a consequence, given a chromatography setup, the sequence
+of a peptide should provide virtually all the information needed to
+predict its retention time.
 
@@
 information about the position of each amino acid in the
-sequence. However, since shuffled versions of the same peptide
+sequence. However, because shuffled versions of the same peptide
 sequence had previously been separated on a RPLC column, the author
@@
 located at different positions in a sequence \cite{Houghten1987}. In
-this context, the authors speculated that different retention
+that context, the authors speculated that different retention
 coefficients would be needed for each position in a peptide
-sequence. Although the exact chemical mechanisms underlying this
+sequence. Although the exact chemical mechanisms underlying that
 observation are still not fully understood, a number of studies have
@@
 features that are used as an input for final retention time
-estimates. We will come back to this in section \ref{sec:ML}.
-
-% - sum of retention coefficients (Meek reference 22) - different RT
-% coefficients and the observation that this coefficients depend on
-% the chromatography system (Browne ref 24) - investogate the
-% contribution of individual amino acids (Guo 26,27) - all the work
-% above was done on synthetic peptides os simple samples. Then RT
-% prediction for complex sa% mples was proposed by (Palmblad 28-30) -
-% important the exact position of each aa (pref 27) - additional rules
-% included peptide length (ref 32), position of K and R residues [ref
-% 33], and in the e% nd that the position of each amino acid at each
-% position is important [ref 34] - amphipathicity [ref 35] - SSRCalc
-% [ref 36-39]
+estimates. We will describe such methods more in detail in section
+\ref{sec:ML}.
 
@@
 with known retention times, a strategy to normalize elution times was
-developed. This allowed pooling of peptides identified in more than
-12,000 shotgun experiments. The trained algorithm reported better
-performance for the 1303 peptides on which it was evaluated, than its
-predecessor.
+developed. That strategy allowed pooling of peptides identified in
+more than 12,000 shotgun experiments. The trained algorithm reported
+better performance for the 1303 peptides on which it was evaluated,
+than its predecessor.
 
@@
 peptides required, which made the algorithm difficult to retrain for
-other chromatographic conditions. This was a serious limitation in the
-context of RPLC separation, where a broad variety of chromatographic
-setups is commonly used across research laboratories. To address this,
-Klammer {\em et al.} proposed a predictor based on support vector
-machines (SVMs) \cite{klammer2007improving}. Their method required
-fewer training peptides and could thus be easily adapted to new
-chromatographic conditions, but yielded lower prediction accuracy.
-Subsequently, a more elaborate SVM-based method, {\sc RTPredict}, that
-uses a customized kernel function was proposed \cite{rtpredict,
-  rtpredictImproved}. This approach provided an algorithm that
+other chromatographic conditions. This requirement was a serious
+limitation in the context of RPLC separation, where a broad variety of
+chromatographic setups is commonly used across research
+laboratories. To address that lack of flexibility, Klammer {\em et
+  al.} proposed a predictor based on support vector machines (SVMs)
+\cite{klammer2007improving}. Their method required fewer training
+peptides and could thus be easily adapted to new chromatographic
+conditions, but yielded lower prediction accuracy.  Subsequently, a
+more elaborate SVM-based method, {\sc RTPredict}, that uses a
+customized kernel function was proposed \cite{rtpredict,
+  rtpredictImproved}. That approach provided an algorithm that
 required as little as 50 peptides for training.
@@
 helicity, and the number of occurrences of the 25\% most or least
-hydrophobic amino acids. This ensured that the algorithm could be
-applied to different chromatography setups with minimal loss in
+hydrophobic amino acids. This scheme ensured that the algorithm could
+be applied to different chromatography setups with minimal loss in
 performance. For cases when little training data was available, {\sc
@@
 
-To address this, in Moruz {\em et al.} \cite{elude2} extended the
-first version of their algorithm {\sc ELUDE} to encompass modified
-peptides as well. This was straightforward to achieve once the authors
-observed that many of the predictive features of their method were all
-functions of the peptide's amino acid sequence and a retention
-index. As such retention indices are estimated for each dataset at
-hand, at the beginning of the training procedure, they can trivially
-be extended beyond the standard amino acids. When the authors used but
-the features based on their retention time index, they obtained a
+To address the limitations of the previous studies, Moruz {\em et al.}
+\cite{elude2} extended the first version of their algorithm {\sc
+  ELUDE} to encompass modified peptides as well. The inclusion of such
+peptides was straightforward to achieve once the authors observed that
+many of the predictive features of their method were all functions of
+the peptide's amino acid sequence and a retention index. As such
+retention indices are estimated for each dataset at hand, at the
+beginning of the training procedure, they can trivially be extended
+beyond the standard amino acids. When the authors used but the
+features based on their retention time index, they obtained a
 negligible loss in accuracy compared to using all their original
-features for unmodified peptides. Since this index was trained for
+features for unmodified peptides. Beacause this index was trained for
 each dataset at hand, the algorithm became fully trainable for any
@@
 differentiated between correct and incorrect peptide identifications
-\cite{Strittmatter2004}. Following this approach, they reported
+\cite{Strittmatter2004}. Following that approach, they reported
 $\sim$9\% more confident peptide identifications in a typical shotgun
@@
 The idea of identifying proteins based on entire peptides, rather than
-peptide fragments, is not new. In fact, this was the principle
-underlying peptide mass fingerprinting, one of the first methods used
-to identify proteins via mass spectrometry. However, when introduced
-in the 90's, mass fingerprinting could not be applied to unequivocally
-identify proteins in complex biological samples. Consequently, shotgun
-proteomics has gained popularity instead. Nevertheless, with great
-developments in instrumentation leading to sub-ppm mass accuracy, the
-possibility of using mass fingerprinting-like methods came back to
-attention~\cite{iamt1}. The mentioned study showed how improved mass
-accuracy and retention time predictions enable the use of unfragmented
-ions even in the context of complex peptide mixtures.
+peptide fragments, is not new. In fact, that strategy was the
+principle underlying peptide mass fingerprinting, one of the first
+methods used to identify proteins via mass spectrometry. However, when
+introduced in the 90's, mass fingerprinting could not be applied to
+unequivocally identify proteins in complex biological
+samples. Consequently, shotgun proteomics has gained popularity
+instead. Nevertheless, with great developments in instrumentation
+leading to sub-ppm mass accuracy, the possibility of using mass
+fingerprinting-like methods came back to attention~\cite{iamt1}. The
+mentioned study showed how improved mass accuracy and retention time
+predictions enable the use of unfragmented ions even in the context of
+complex peptide mixtures.
 
@@
 In addition, predicted retention times are useful investigating the
-elution properties of the peptides.  To illustrate this, in
+elution properties of the peptides.  To illustrate this ability, in
 \cite{elude1} the retention time of all the theoretical peptides from
@@
 a majority of the peptides eluting in only a small portion of the
-chromatography run. This suggested that the linear gradient typically
-used in liquid chromatography are sub-optimal, and that one should be
-able to replace such gradients with nonlinear functions that produce
-even distributions of the peptides across runs.
+chromatography run. This behavior suggested that the linear gradient
+typically used in liquid chromatography are sub-optimal, and that one
+should be able to replace such gradients with nonlinear functions that
+produce even distributions of the peptides across runs.
 
@@
 practitioner, a user-friendly graphical user interface was implemented
-\cite{gradopt2}.  Seen from a general perspective, this showed that
-the gradient function is yet another parameter that researchers can
-alter in their experiments, opening up a whole new range of potential
-applications.
+\cite{gradopt2}.  Seen from a general perspective, these studies
+showed that the gradient function is yet another parameter that
+researchers can alter in their experiments, opening up a whole new
+range of potential applications.
 
@@
 \cite{liu2009learning}, which we feel is particularly suitable for
-this problem. MLR models can be trained to determine which of two data
+that problem. MLR models can be trained to determine which of two data
 points should be better ranked.
@@
 using restricted multiple layered Boltzmann machines (RBM)
-\cite{salakhutdinov2009deep} or deep neural networks. This follows the
-observation that the performance of predictors is often enhanced if
-one can structure the prediction into multiple layers. The drawback of
-such models is that they require large training sets. Given the very
-large sets of peptides with retention times that are assembled in
-various repositories, such as Peptide Atlas
+\cite{salakhutdinov2009deep} or deep neural networks. This remark
+follows the observation that the performance of predictors is often
+enhanced if one can structure the prediction into multiple layers. The
+drawback of such models is that they require large training
+sets. Given the very large sets of peptides with retention times that
+are assembled in various repositories, such as Peptide Atlas
 \cite{desiere2006peptideatlas} and PRIDE \cite{martens2005pride}, such
@@
 
-One way to determine a predictable separation technique's usefulness is
+One way to measure a predictable separation technique's usefulness is
 to estimate its ability to remove candidate peptides of any detected
-feature. This can be measured by a false positive ratio (FPR), also
-known as the method's fall-out. The FPR is of course a function of the
-selected thresholds, but if we select a threshold where $\sim$95\% of
-the known examples get correctly included, what FPR do we find at such
-level?  Current retention time prediction algorithms have an FPR of
-about $\frac{1}{5}$, meaning that it can remove about four out of five
-candidate peptides for a given feature. This could be compared to the
-FPR of about $\frac{1}{200}$ for predictions of high-resolution
-isoelectric focusing experiments~\cite{branca2014hirief}. The
-retention time prediction has to improve quite a bit to come to level
-with the fall-out of isoelectric focusing, however, almost all mass
-spectrometry-based proteomics use reversed phase chromatography, while
-very few studies are using iso-electric focusing.
+feature. This property can be measured by a false positive ratio
+(FPR), also known as the method's fall-out. The FPR is a function of
+the selected thresholds, but if we select a threshold where $\sim$95\%
+of the known examples get correctly included, what FPR do we find at
+such level?  Current retention time prediction algorithms have an FPR
+of about $\frac{1}{5}$, meaning that it can remove about four out of
+five candidate peptides for a given feature. This rate could be
+compared to the FPR of about $\frac{1}{200}$ for predictions of
+high-resolution isoelectric focusing
+experiments~\cite{branca2014hirief}. The retention time prediction has
+to improve quite a bit to come to level with the fall-out of
+isoelectric focusing, however, almost all mass spectrometry-based
+proteomics use reversed phase chromatography, while very few studies
+are using isoelectric focusing. As, an additional comparison is the
+selection process based on the prediction of peptides' masses and
+determination using mass spectrometry. A typical 10 ppm fully tryptic
+search have on average $\sim 150$ candidate peptides per fragment
+spectrum, out of the total of $\sim 4\cdot 10^6$ fully tryptic
+peptides in the swissprot part of UniProt (including oxidation and
+N-terminal acetylations), indicating that the FPR of the ms1-mass
+determination/prediction is about $4\cdot 10^{-5}$. Neither
+chromatographic retention time prediction, nor isoelectric focusing,
+is currently near such performance.
 
-% An estimate of a FPR for MS1's would be great here.
-
-Moreover, targeted proteomics experiments, such as Selected Reaction
-Monitoring (SRM), are dependent on the predictability of a peptide's
-behavior. In such experiments the retention times and the suitable
-transitions are needed as input to the analysis. Frequently, such
-information is derived from previous shotgun analysis. As the capacity
-of targeted proteomics experiments expands by the introduction of the
-more comprehensive data independent acquisition techniques
+Targeted proteomics experiments, such as Selected Reaction Monitoring
+(SRM), are dependent on the predictability of a peptide's behavior. In
+such experiments the retention times and the suitable transitions are
+needed as input to the analysis. Frequently, such information is
+derived from previous shotgun analysis. As the capacity of targeted
+proteomics experiments expands by the introduction of the more
+comprehensive data independent acquisition techniques
 \cite{Venable2004} (a.k.a. SWATH analysis), such pre-identification of

### 2015-12-22 retention.tex
x

@@
 \setlength{\parskip}{0.15cm}
+
+\begin{abstract}
+Most methods for interpreting data from shotgun proteomics experiments
+are to large degree dependent on being able to predict properties of
+peptide-ions. Often such predicted properties are limited to molecular
+mass and fragment spectra, but here we put focus on a perhaps
+underutilized property, a peptide's chromatographic retention time. We
+review a couple of different principles of retention time prediction,
+and their applications within computational proteomics.
+\end{abstract}
+
+Keywords: Bioinformatics; Chromatography; Regression analysis; Mass spectrometry; Peptides
 
